### 2017-04-13 Papers/DraftOutline_SSCCM.tex
first commit of datafiles

\documentclass[]{article}

\begin{document}

\title{Site-Heterogeneous Character Change Models for Morphology}
\author{Wright, Pett, Students, Heath}
\date{Today}
\maketitle

\section{Introduction}
\subsection{Bayesian Modeling of Morphology}
	Recent interest in using Bayesian methods to model morphological evolution to estimate phylogenetic trees has spurred many studies into how well the existing toolkit performs, particularly in relation to parsimony methods.
	However, there has been little work expanding that toolkit. 
	The most common way to perform likelihood or Bayesian phylogenetic analysis using morphological data is by applying a model of morphological evolution called the Mk model.
	Many workers using morphological data have raised questions about the realism of this model.
	In this paper, we will discuss methodological advancements aimed at improving the realism of the model. \par
	The Mk model was introduced by Lewis in 2001.
	As a generalization of the Jukes-Cantor model, it makes the same set of assumptions: that change between any two states is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} states.
	The model also makes the Markovian assumption that a character can change instantaneously, regardless of previous states. 
	Any one of these assumptions may strike a reader as problematic for their particular dataset. \par
	Heterogeneity in the evolutionary process is difficult to accommodate in morphological data. 
	Making concrete \textit{a priori} statements about the relative probabilities of transitions between character states in a morphological character matrix is challenging. 
	Unlike a matrix of nucleotide characters, a `0' character at one state may mean that the trait never evolved in one lineage, but was lost in another.
	This lack of common meaning complicates the interpretation of how change occurs over time.
	Likewise, the magnitude of change from one state to another may be different between characters, even if the characters mean the same thing (i.e. `0' representing loss and `1' representing gain).
	This has complicated the ability to specify $\alpha$ parameters to the Q-matrix \textit{a priori}.
	This is a stark contrast to molecular data, in which base pairs or nucleotide residues are generally assumed to have similar properties across an alignment. \par
Prior work extending the Mk model has focused on relaxing the assumption of equal character frequencies at stationarity.
MrBayes implemented a parameter called the Symmetric Dirichlet Hyperprior, which allowed users to place a prior on state frequencies. 
The probability of observing a change in a character is dependent not only on the probability of change from this character to another, but on the frequency of that character.
Even if a character has a low probability of changing, change in that character may still be observed many times if the stationary frequency of that character is high (i.e., the character is observed many times). 
Likewise, a highly-probable change may be seen relatively rarely if the starting character is observed rarely. 
Therefore, relaxing the assumption of equal stationary frequencies has been a way of changing the probabilities of observing different changes without making strong \textit{a priori} statements about transition probabilities. \par
This analysis was implemented as a model extension in MrBayes.
In the case of binary data, this was a Beta prior (the Dirichlet is a generalization of the Beta distribution), which operated in a fashion similar to gamma-distributed rate variation: the distribution is discretized into a user-specified number of categories, the median forward and backward transition rate are calculated for each category, and the likelihood of the character is calculated over each category and summed to form the total likelihood. 
For binary data, the state frequencies are integrated out of the likelihood, making the symmetric Dirichlet a prior.
For multistate data, the state frequencies are drawn from a Dirichlet distribution.
The parameter was referred to as a hyperprior because users can place a distribution on the parameter to the Beta distribution, and because the state frequencies are not integrated out of the likelihood function for multistate data. 
As implemented in MrBayes, the Beta distribution was assumed to be symmetric, meaning that if there were characters for which the stationary frequency of 0 is high, there would also be corresponding characters for which the stationary frequency of 1 is high.\par
This is a very useful model extension for morphological evolution.
There are clear contexts to improve this concept by borrowing from the molecular literature.
The CAT model of Lartillot represents one way forward. 
This model, implemented in PhyloBayes, uses a Dirichlet process to assign individual sites to categories, which differ in their stationary frequencies. 
In this model, the number of such categories that describe the data is a free parameter. 
The use of a Dirichlet prior allows for flexibility in the number of states, as opposed to a Beta prior which is limited to binary data.
However, many datasets that have been analyzed under the CAT model are phylogenomic datasets of thousands of loci, as compared to small morphology datasets.\par
In RevBayes, we have implemented a number of useful extensions to the Dirichlet prior of RevBayes, and a constrained version of the CAT model. 
Our extension to the symmetric Beta prior allows for the Beta distribution to be asymmetrical.
This removes the need for there to be equal numbers of characters with high stationary frequencies for a character state as there are characters with low stationary frequencies for that same character state.
Our CAT-like model (hereafter Site-Heterogeneous Discrete Morphology (SDHM) model) is a finite mixture model that allows characters to fit into a user-defined number of character categories that differ in their stationary frequencies. 
The bins draw their stationary frequencies from a Dirichlet to accommodate multistate data. 
\subsection{Morphological Phylogeny of the Formicidae}
To test the efficacy of our analytical techniques, we use an ant dataset created from datasets from Barden and Grimaldi and one from Keller.
Ants are ecologically crucial organisms as both interacting partners for a variety of plants and animals, and as shapers of the ecosystem via soil cycling and nest building.
As such, they have attracted much work in the world of molecular systematics.
Ants have also have a rich fossil record, with most subfamilies being represented. \par
The monophyly of major subfamilies has been consistently supported by most molecular and morphological studies, with the exception of the Cerapachyinae. 
Molecular systematics had originally shaken up the ant tree of life considerably, breaking up clades considered to be monophyletic based on morphological work.
However, recent family-wide molecular systematic studies provided support for previous morphological work, supporting a clade of subfamilies Ectatomminae and Heteroponerinae, with Myrmicinae as sister to this clade. \par
Excellent morphological matrices on the ingroup of the Formicidae have been available since the early '90s. 
Recent morphological matrices have expanded the sampling of ants from in the large Ponerine family, which was previously undersampled.
Sampling of fossil ants was also expanded to include specimens from the Cretaceous.
These ants were concluded to be stem lineages, and some could not be concretely placed within lineages. \par
In this study, we combine the extant matrices of Keller, and the extinct-extant matrix of Barden and Grimaldi to expand the taxon and character sampling.  
This dataset has interesting properties.
Because there are stem ant lineages represented, there are taxa that have characteristics that are lost after the divergence of the stem lineages.
This extremely one-sided loss structure violates the Mk model assumption of equal transition rates. 
We would expect for these characters to be better modeled by a model that can accommodate asymmetrical transition rates.
Some apomorphies of the ant group are also gained after the divergence of the stem lineages, which also violates the assumption of equal change probabilities.
Because of these model violations, this dataset is an excellent test case for models that relax assumptions of the Mk model. \par
We use Bayes Factor model selection to assess the fit of these relaxed models to the data.
Using this dataset, we strongly support that the use of models that relax key assumptions of the Mk model can greatly improve the fit of the model to the data.
Some stuff about trees and parameters!

\section{Methods}

	a) Bayes Factor model fitting
	b) TreeSet viz showing how use of the asymetric beta changes our pattern of exploration of treespace

Results

	a) Best fit model?
	b) Tree comparisons, esp. branch lengths. No asc. bias correction + misspecified model, no correction + misspecified model, correction + correct model, correction + misspecified model 
	c) Tree set visualization comparisons

Discussion

	a) How has the modeling of morphology changed since the Mk was proposed by Lewis? What aspects are comparable to how we model DNA? Which are not? What are potentially interesting future directions? 
	b) General discussion of usage of treeset visualization
	c) something else.

\end{document}

### 2017-04-13 Papers/DraftOutline_ascBias.tex
first commit of datafiles

\documentclass[]{article}

\begin{document}

\title{Comment on Missing Data, Model Misspecification and Ascertainment Bias}
\author{Wright, Pett, Holder, Heath}
\date{Today}
\maketitle

\section{Introduction}
	For decades, phylogenetic inference from discrete morphological data has primarily been performed using the maximum parsimony optimality criterion. 
	Parsimony is conceptually easy to understand: a phylogenetic tree is proposed to explain a matrix of morphological characters.
	That tree is scored according to how many changes in the character matrix are required to explain it.
	Conceptual extensions to parsimony, such as weighting different types of character changes also became common. 
	However, in recent years, there has been an increased interest in using Bayesian methods for the inference of phylogeny from morphological data. 
	\par
	This interest has been driven, in part, by acknowledgement that incorporating more fossil data improves estimates of topology and divergence date analyses in combined molecular-morphological analyses.
	Since molecular systematists primarily work with likelihood and Bayesian analyses, more researchers began to look for ways to incorporate fossils into the analytical frameworks with which they were familiar.
	This interest also came from researchers working with comparative methods.
	Without the inclusion of fossils, which are our only direct observations of the past, we may miss historical evolutionary dynamics.
	For paleontologists, making use of the toolkit offered by phylogenetic comparative methods means estimating trees that have rate-based branch lengths, typically expressed in expected changes per character.
	This type of branch length is estimated by Bayesian and likelihood analyses, but not parsimony analyses. 
	Bayesian methods also calculate distributions of parameters and trees, rather than point estimates.
	This leads to a more natural incorporation of uncertainty. \par
	This renewed interest has lead to a flurry of papers examining Bayesian approaches to morphological phylogenetics.
	However, many of the papers proposing new methods for incorporation of morphological data appeared in journals whose readership is primarily molecular. 
	Here, we review some of the challenges of building Bayesian phylogenetic trees from morphological data.
	\subsection{What is a Bayesian analysis?}
	In any statistical approach to phylogenetics, evidence in favor of a particular tree topology is summarized by the likelihood function $P(D\mid T,\theta)$, which is the probability of observing the character data $D$, given a specific tree topology $T$ (and other model parameter values $\theta$).
	In a maximum likelihood analysis, the goal is to identify the values of $T,\theta$ that maximize the likelihood function.
	Uncertainty in this maximum likelihood estimate is typically evaluated using statistical resampling techniques, such as bootstrap or jackknife methods.

	In a Bayesian analysis, the goal is to infer the posterior probability $P(T,\theta\mid D)$.
	The posterior probability has a natural and intuitive interpretation as the degree of belief in a particular tree (and parameters) after observing data $D$.
	In addition, all uncertainty in $T,\theta$ is described by their posterior distribution.
	The posterior distribution of $T,\theta$ can be computed from the likelihood using Bayes theorem $$P(T,\theta\mid D) = \frac{P(D\mid T,\theta)P(T,\theta)}{P(D)}$$

	The term $P(T,\theta)$ represents the \emph{prior distribution}, or our degree of belief in $T,\theta$ before observing any data.
	While it may seem counter-intuitive to specify such prior beliefs, a prior distribution is actually implicit in any maximum likelihood analysis as well.
	These assumptions are made explicit in a Bayesian analysis through specification of the prior distribution.
	Furthermore, by explicitly specifying the prior, it is possible to accommodate any prior knowledge/constraints we may have for the tree topology or model parameters.

	The constant term $P(D)$ represents the \emph{marginal likelihood} of the data, or the expected value of the likelihood averaged over the prior distribution of all possible parameter values.
	The computation of this term is computationally intractable for most models, and therefore specialized numerical methods are required to compute the posterior probability.
	Chief among these are Markov Chain Monte Carlo (MCMC) methods, which produce a sample from the posterior distribution of trees and parameters by following a random walk through parameter space.
	The key feature of this random walk is that its movements are directed from one state to another in proportion to the ratio of the posterior probability computed at each state.
	Conveniently, because the constant term $P(D)$ cancels out in this ratio, it is therefore possible to sample from the posterior distribution using MCMC without ever computing the marginal likelihood.
	The end result is a sample from the posterior distribution, which completely describes our knowledge and uncertainty of the tree and model parameters after observing data $D$.

	\subsection{How have Bayesian methods been applied to the fossil record?}
	In 2001, Paul Lewis proposed a model called the Mk model, which stood for Markov \textit{k}-states model.
	This model was a generalization of the Jukes-Cantor model for molecular sequence evolution.
	As such, it makes  very few assumptions about the data.
	The Mk model makes the standard Markovian assumption that a character can change instantaneously from one instant on a branch to the next.
	Being a generalization of the Jukes-Cantor model of sequence evolution, the Mk model makes the same assumptions: chief among them that a character in any of the \textit{k} states specified by the user can transition to any of the other states with equal probability.
	This is a very simple model, and the assumptions are not dissimilar to the assumptions made to an equal-weights parsimony step matrix. \par
	The Mk model has been shown by mathematicians to be statistically consistent, meaning that as the amount of data provided approach infinite, the model will converge to the true answer.
	By contrast, in many regions of parameter space, parsimony is shown to be positively misleading, meaning that wrong answers will be supported with higher probability as data are added.
	The increased interest in Bayesian phylogenetics for the fossil record has resulted in several simulation studies looking into the efficacy of the model.
	Ultimately, these studies conclude that using Bayesian methods to look at the fossil record often presents an increase in accuracy, while incorporating uncertainty more naturally. \par

	\subsection{What are specific concerns for modeling morphological data using Bayesian methods?}
	Nucleotide characters have different properties than morphological characters.
	Due to the nature of collecting morphological data, there are special considerations that must be taken into account to model characters.
	One major consideration is that researchers have generally only collected characters that vary within their clade of interest.
	This means that certain subsets of characters, namely those which do not vary, are systematically excluded. 
	Therefore, we must correct for this bias, called ascertainment bias, when setting up our models. \par
	Morphological characters are also coded arbitrarily, but non-randomly.
	Nucleotide, amino acid and protein data all have codings that correspond directly to a biochemical meaning.
	A character state in morphology does not have these same explicit meanings.
	A human makes the determination, according to their notions of character evolution, what to call each state.
	This property of human bias in character coding also raises challenges for modeling the data. \par
	Below, we will discuss in more depth each of these two issues.

	\subsubsection{What is ascertainment bias? Why do we correct for it?}

	Ascertainment bias refers, broadly, to the systematic under- or oversampling of classes of characters in an analysis.
	In phylogenetics involving discrete morphological characters, this typically means certain character types have been excluded.
	Because trees have generally been estimated using parsimony methods, workers in paleontology typically have not collected characters that are not \textit{parsimony-informative}.
	A parsimony-informative character is one that can be used to favor one set of trees over another.
	Under this definition, invariant characters and autapomorphies are not parsimony-informative. 
	However, in a Bayesian context, these characters are still useful for estimating rate of evolution parameters.
	Their exclusion can lead to serious biases in branch lengths, and as branch lengths and topology are co-estimated, can even bias topology. \par
	When Lewis first introduced the Mk model, he also introduced a correction for this bias.
	This correction simulated invariant dummy characters, and calculated the likelihood of them along the tree. 
	This likelihood was used to normalize the likelihood of the rest of the data. \par
	In this paper, we apply a dynamic programming
	approach to analytically derive the same likelihood.
	The dynamic likelihood approach works by mechanisms that WILL will explain. \par
	It should be noted that branch lengths in a morphological analysis, even when appropriate corrections are used, are still sensitive to the prior distribution chosen. 
	In response to this sensitivity, we describe below a hierarchical prior distribution on branch lengths that allows for signal to be extracted from the data more efficiently. \par

	\subsubsection{What does arbitrary character coding mean for how we model morphology?}

	Morphological data are coded in a way that is \textit{arbitrary}, but meaningful and non-random.
	The character codings are arbitrary in the sense that they do not have universal meaning.
	For example, when we collect nucleotide data, are collected, an `A' state has explicit meaning: it represents the nucleotide adenine, with all of adenine's biochemical properties.
	These biochemical properties allow us to make inferences about the types of changes that are likely to be observed over evolutionary time.
	For example, adenine is a purine nucleotide base, and we are more likely to observe changes from adenine to the other purine (guanine) than to the pyrimadine bases (cytosine and tyrosine). \par
	Morphological characters lack these sorts of objective characterizations.
	Different workers can code the same morphological character as binary, or as multistate. 
	However, some meanings have arisen by convention.
	The character state `0' often refers to absence.
	Different researchers, may, however use the `0' character to refer only to characters that have not evolved in a lineage, while others may use it to refer both to the character not having evolved as well as the character having been lost. \par
	The Mk model, as proposed by Lewis, made the assumption that character state frequencies were equal, and that all characters had the same frequencies.
	In the software MrBayes, Huelsenbeck and Ronquist provided an option to relax the equal frequencies assumption.
	In their method, researchers could place a symmetrical Beta prior on state frequencies, which would allow them to vary across characters. 
	The Beta distribution would be discretized into a user-defined number of categories.
	The value state frequencies for a given category would then be equal to the mean of that category. 
	This means that for a given character, the state frequencies could be unequal.
	The likelihood of observing a specific character state change is a product of both the transition rate and the frequency of the character state that is changing.
	For example, it might be extremely likely to change from having a a limb be partially articulated to being full articulated.
	But if few taxa with partially articulated limbs are in the sample of interest, this change will be seldom observed. 
	A low-probability transition, by contrast, may be observed many times, if the character frequency of the originating state is common. \par
	While this method allows for individual characters to have differential frequencies, altering the frequency with which a particular change is observed, the distribution is still symmetrical.
	If the Beta distribution was discretized into four categories, then there would be four possible sets of state frequencies. 
	Because of the property of symmetry, if one category had a very low frequency, this would be balanced by a high frequency category.
	Wright, Pett et al. implemented an asymmetrical discretized Beta to allow for data where this assumption is likely to be violated.\par
	Because this model uses a Beta distribution, it is not coherent for multistate data.
	Wright, Pett et al. also introduced a mixture model to allow state frequency heterogeneity that works with multistate data.
	The Site-Heterogeneous Discrete Morphology model (SHDM) uses a Dirichlet distribution to draw state frequencies from an \textit{N}-dimnesional space, in which \textit{N} is the number of character states. 
	This model is called a mixture model because it is assumed that there are categories within the data that should be described with their own state frequencies. 
	These categories are not specified \textit{a priori}, but are rather derived analytically. 
	This model is a generalization for morphological data of the CAT model described by Lartillot. 
	In effect, SHDM allows for state frequency heterogeneity models to expand to multi-state data.
	\par

 \subsection{Advances in Modeling Paleobiological Data}
	Here, we perform multiple analyses that address the aforementioned challenges to modeling discrete morphological data in a Bayesian context in the software package RevBayes. 
	Using an example dataset from the Formicidae, we demonstrate the utility of several new approaches to modeling arbitrary character coding and ascertainment bias. 
	We also use simulated data to demonstrate the efficacy of new corrections for ascertainment bias. 
	Finally, we demonstrate the effectiveness of a hierarchical framework for modeling the branch lengths of the tree. \par

\section{Methods}

\subsection{Datasets}

We used two datasets for this paper, both of which come from previously-published papers.
One is from Barden and Grimaldi.
This dataset contains 41 extinct ant taxa, with 41 characters.
26 characters were binary, and 15 were multi-state.
The original dataset contained one character with more than 10 states.
RevBayes, the software in which the phylogenetic estimations were performed, cannot use characters with more than ten states.
This character was pruned from the final dataset. \par
A second set of simulated datasets was used to test the ascertainment bias corrections, which were unpublished for morphological data.
These 350-character datasets are from Wright and Hillis, and were previously used to test the efficacy of the ascertainment bias corrections in MrBayes. 
They were simulated along a 76-tip tree from Pyron, under the Mk model. \par

\subsection{Ascertainment Bias}

To test the efficacy of the ascertainment bias correction, we performed several comparisons with simulated data.
First, we estimated the likelihood of the tree with unfiltered data (data containing parsimony-informative, invariant and non-parsimony-informative variable characters) using RevBayes and MrBayes.
Then, in both RevBayes and MrBayes, we estimated the likelihood of the datasets after having filtered out parsimony non-informative variation, using the ascertainment bias corrections implemented in those programs.
RevBayes and MrBayes should return the same likelihoods for the datasets, as they are calculating the same likelihood correction. \par

\subsection{Character Coding}

In RevBayes, we used the Formicidae dataset to test the Mk model against two more complex models for morphology data. 
All test datasets were run for 3 million generations, and used Gamma-distributed rate variation (GDVR). 
For more precise model-testing, we also performed stepping-stone analysis, which computes the likelihood of a model more accurately to enable better comparisons of models. \par
The first model tested is the Mk model as formulated by Lewis (2001).
The second is a simple model for parameterizing state frequency heterogeneity in binary data. 
This model is a generalization of the discretized Beta distribution described in the section \textbf{1.3.2}. 
The final model is the SHDM described in \textbf{1.3.2}. 
We used this model for both the binary and the combined binary and multistate datasets. 
\par

\subsection{Branch Lengths}

Branch lengths in Bayesian morphological analyses have not been addressed extensively in the literature.
MrBayes, by default used an exponential distribution centered at 10 to describe branch lengths.
In RevBayes, we varied this parameter.
We used an exponential centered at 10, one centered at 5, and one in which the mean of the exponential distribution was drawn from a distribution.
The latter technique is called a hierarchical prior, and allows the shape of the prior distribution to vary according to the data. \par

\section{Results}

	a) Likelihood scores comparison - Fig 1 for asc bias
	b) Tree comparisons, tree lengths. No asc. bias correction + misspecified model, no correction + misspecified model, correction + correct model, correction + misspecified model Fig 2a
	c) Show that the dynamic programming and dummy char approaches work about the same. Fig 1b
	d) L comparisons for character coding
	e) Tree comparisons for no\_corr, Mk, Mk+disc, Mk+cat
	f) branch lengths

Discussion

	a) How has the modeling of morphology changed since the Mk was proposed by Lewis?
	b) Ascertainment bias: how does missing data interact with ascertainment bias to make corrections important
	c) What does this mean for data collection? And for dataset reuse?

\end{document}
\end{document}

### 2017-04-13 Papers/Intro_short.tex
first commit of datafiles

\documentclass[]{article}
\usepackage[affil-it]{authblk}

\begin{document}

\title{Editor's Note on `Putting Fossils in Trees' Special Issue}
\author[1,2]{April M. Wright}
\affil[1]{Ecology, Evolution and Organismal Biology, Iowa State University, Ames, IA, 50010}
\affil[2]{The Field Museum, Chicago, IL, 60605 \\ wright.aprilm@gmail.com}

\maketitle

This special issue has its origins with a 2014 symposium organized by myself and my co-editors David W. Bapst, Graeme T. Lloyd and Nicholas J. Matzke at the  Society for Vertebrate Paleontology meetings in Berlin, Germany.  In the two years prior to 2014, there had been several interesting and important papers published \cite{stadler2010sampling} 
\cite{ronquist2012total} \cite{bapst2013stochastic}\cite{heath2014fossilized}   about estimating phylogenetic trees incorporating fossil taxa, and particularly involving Bayesian analyses. Some of these papers were written from a neontological perspective, and appeared in journals not often read by paleontologists. We convened our session with a simple idea: to get different types of researchers from both neontology and paleontology together to share work and ideas concerning phylogenetic analyses in paleontology. The session's success inspired my co-editors and I  to propose a Special Feature for Biology Letters, to which we invited the session participants and other contributors. The Biology Letters short format is perfect for focused vignettes about the actual practice of phylogenetics with fossils. \par

For years, those working primarily with fossil data, and those working primarily with molecular data, took different approaches to both building phylogenetic trees, and scaling them to absolute time. Paleontologists have often applied a maximum parsimony paradigm under which potential trees are scored according to how many character changes are required in the dataset to produce that tree. The tree implying the fewest changes being ultimately preferred, returning either the single most parsimonious tree or a sample of equally parsimonious trees. Neontologists, working primarily with molecular data, more often apply likelihood-based methods, including Bayesian methods, which evaluate phylogenetic trees by applying a model of evolution to the data, and calculating the likelihood of the data given the tree and the model. Maximum likelihood methods return a point estimate of the phylogeny with branch lengths equal to the expected number of substitutions per site along that branch. Bayesian methods allow for researchers to quantify and incorporate their existing knowledge about phylogenetic relationships and model parameters using probability distributions called priors. Under Bayesian methods, a sample of trees, allowing a more natural incorporation of uncertainty in phylogenetic estimation. \par

Communication between palaeontological and neontological workers has been limited. No matter the methodology for estimating them, phylogenetic trees have become crucial to comparative biology. Phylogenetic trees have been used to formulate, give evidence in favor of or against hypotheses about the evolutionary history of groups and traits. Because the field of phylogenetics is both crucial and rapidly-changing, we sought to incorporate a diverse cross-section of the approaches biologists use to incorporate fossils in trees, and how those trees are eventually used to inform the study of evolution. \par

The bulk of these papers have focused on the estimation of phylogenetic trees. Building a quality phylogenetic dataset remains a challenge for researchers; Guillerme and Cooper\cite{guillerme2016assessment} address this issue in their study on availability of both fossil and extant data in mammals. This paper takes a new look at the very old question of if and how missing data is injurious to phylogenetic inference, concluding that the distribution of missing data is more important than the ultimate amount. O'Reilly and Donoghue \cite{o2016tips} take a broad look at different paradigms for dating phylogenetic trees, using fossils either as tips in the analysis or as node calibrations, ultimately concluding that these approaches can be used concurrently. Matzke and Wright \cite{matzke2016inferring} evaluate the importance of the assumptions made in Bayesian divergence dating analyses, and demonstrate how violations of these assumptions can result in poor inferences. Together, these papers address some of the aspects of building and dating phylogenetic trees incorporating fossils that may be opaque to researchers. \par

Some papers built trees as a way to understand the biology of a specific clade, or to address a question about how traits have evolved, and how methodological advances can allow us to explore these questions more thoroughly. Researchers in this issue tackled two phylogenetic questions that hold not just the interest of scientists, but the general public. Lloyd et al. \cite{lloyd2016probabilistic} use the fossil record of dinosaurs in conjunction with model-based methods for a posteriori time-scaling to evaluate when the bird clade, and flight, originated. Lee \cite{lee2016multiple}, seeking to understand the timing of mammal evolution, implemented automated approaches for discovering which parts of the dataset evolve under potentially different evolutionary rates. He  concludes that automated approaches can be useful for understanding the biology that underlies phylogeny.\par

Lastly, some of the papers in this issue explored the downstream use of phylogenetic trees in trait-based or biogeographic analysis. Halliday and Goswami \cite{halliday2016impact}, and Bapst et al. \cite{bapst2016topology} both evaluate multiple phylogenetic trees in phylogenetic comparative analyses. Halliday and Goswami built both parsimony and Bayesian phylogenies of placental mammals to look at body-size evolution. Bapst et al. perform a similar analysis in birds. Both papers ultimately conclude that careful attention to the assumptions underlying the production of dated phylogenies is crucial, and that the inferential method chosen can greatly affects the results of comparative analyses. Gorscak and O'Connor \cite{gorscak2016time} build a Bayesian time-scaled phylogeny to understand the paleobiogeography of titanosaur sauropod dinosaurs, in the process posing several large taxonomic rearrangements. They then use a model selection approach to conclude that titanosaurs were globally distributed, but experienced regionalization as the continents drifted apart in the Cretaceous. Together, these three papers suggest that to learn more about evolutionary patterns via phylogenies with fossil tips, we must closely examine how we reconstruct our dated trees.\par

It has been a pleasure to assemble this collection. I am proud that seven of the papers in this special feature come from early-career researchers, including graduate students. I am particularly thankful to Biology Letters, and editors Surayya Johar and Raminder Suresh for assisting and allowing us to gather a diverse and talented group of researchers to make this special issue. This field has moved faster than anyone could have expected. In a few short years, the models and methods available to analyze combined datasets have expanded greatly. My co-editors and I look forward to seeing how methods for putting fossils in trees will continue to develop into the future.\par

\bibliographystyle{vancouver}
\bibliography{Intro_short}
\end{document}

### 2017-04-13 Papers/coverletter.tex
first commit of datafiles

\documentclass[12pt]{article}
\usepackage[a4paper,margin=1cm,footskip=.5cm]{geometry}
\usepackage[]{authblk}
\usepackage{graphicx}
\usepackage{hanging}
\usepackage{indentfirst}
\usepackage{fancyhdr}
\usepackage{setspace}
\usepackage{multicol}
\usepackage{color}
\usepackage{lipsum}
\usepackage[breaklinks]{hyperref}
\usepackage[all]{hypcap}
\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
%\onehalfspace
\setlength{\evensidemargin}{0in}
\setlength{\oddsidemargin}{0in}
\setlength{\paperheight}{11in}
\setlength{\paperwidth}{8.75in}
\setlength{\tabcolsep}{0in}
\setlength{\textheight}{9in} % This changed from 9in
\setlength{\textwidth}{6.5in}
\setlength{\topmargin}{0.15in}
\setlength{\topskip}{0mm}
\setlength{\voffset}{-0.5in}
\setlength{\parindent}{1cm}
%\setlength{\headheight}{99pt}
\setlength{\headheight}{0.5in}
\parskip = 1mm
\pagestyle{plain}

\definecolor{citescol}{RGB}{73,0,165}
\definecolor{urlscol}{RGB}{3,0,255}
\definecolor{linkscol}{RGB}{187,24,0}
\definecolor{isured}{RGB}{204,0,51}
\definecolor{isukhaki}{RGB}{76,69,43}

\hypersetup{colorlinks=true,linkcolor=linkscol,citecolor=citescol,urlcolor=urlscol}

\lhead{{\textcolor{isured}{\LARGE{\textsc{Iowa State University}}}}\\{\textcolor{isukhaki}{OF SCIENCE AND TECHNOLOGY}}}
\rhead{\scriptsize{\includegraphics[width=0.85in]{./assets/EEOB-eps-converted-to}\\{\textcolor{isured}{Ecology, Evolution, and Organismal Biology}}}}

\begin{document}

Dear Surayya,

Thank you very much for your prompt handling of my manuscript ``Editor's Note on `Putting Fossils in Trees' Special Issue". In this revision, I have addressed the comments submitted by the reviewer.

Thank you for your time, and I look forward to hearing from you.

\vspace{6mm}
\noindent Sincerely,\\
\includegraphics[width=2in]{./assets/Sig2}\\
\noindent Dr. April M. Wright\\
Postdoctoral Researcher\\
Dept. Ecology, Evolution, \& Organismal Biology\\
Iowa State University\\
251 Bessey Hall\\
Ames, IA 50011\\
\href{mailto:wright.aprilm@gmail.com}{{wright.aprilm@gmail.com}}\\
(512) 940-5761

\end{document}

### 2017-04-13 Papers/ed.tex
first commit of datafiles

\documentclass[]{article}

\begin{document}

\title{Student Research Experiences: Training Biodiversity Scientists Through Data Science}
\author{Wright, Pett, Heath}
\date{Today}
\maketitle

\section{Introduction}

Phylogenetic trees are foundational in the study of evolution. 
We see phylogenetic trees in the study of almost all types of evolutionary studies.
From how we apply medicine to choose an influenza vaccine from year to year, to how we understand the evolution of ecologically-influential traits, phylogeny helps us formulate and test hypotheses about the natural world. \par
Phylogenetic trees can be computed in multiple ways.
Workers using morphological data tend to use parsimony to infer phylogenetic trees.
The parsimony criterion is fairly simplistic: a tree is proposed to explain a data matrix provided by the researcher, and the number of changes implied by the tree is scored.
The tree implying the fewest changes is preferred. 
Similarly, researchers working with large datasets (such as genomic data from microbial populations)  often infer trees using neighbor-joining, in which a matrix of the genetic distances between tips is calculated from the data matrix.
Then, the pairs of tips that have the smallest genetic distances are joined together.
\par
However, most workers in molecular systematics, comparative methods and the study of molecular evolution infer phylogenetic trees using maximum likelihood or Bayesian methods.
We will refer to this framework as statistical phylogenetics.
Statistical phylogenetics makes use of the statistical power of parametric statistics to more robustly infer phylogenetic trees.
While these methods have been consistently supported to perform better at estimating phylogenetic trees, they are also harder to teach.
While parsimony and neighbor-joining are straightfoward counting algorithms, statistical phylogenetics involves understanding the biology of the study organisms and having statistical and computational knowledge. 
For this reason, getting students involved in statistical phylogenetics is a challenge, particularly for early career undergraduates. \par

\end{document}

### 2017-04-13 Spring2017/first_script_PM.py

@@ -1,0 +1,33 @@
+import pandas as pd
+import sys as sys
+import os as os
+
+def extract_rows(data, ant_names):
+    '''A function to extract relevant data from dataframe'''
+    new_data = data.loc[ant_names]
+    return(new_data)
+
+def get_names(namefile):
+   '''Read in a file of names'''
+   with open(namefile, 'r') as f:
+       ant_names = f.read().splitlines() 
+   return(ant_names)
+
+def check_dir(path):
+	check=path.split('/')[0]
+	os.listdir(check)
+	if 'output' in os.listdir(check):
+		return
+	else:
+		os.mkdir('%s/output' % check)
+
+if __name__ == '__main__':
+	data = pd.read_csv(sys.argv[1], index_col=0)
+	''' ../../Data/Ants.csv '''
+	namefile = sys.argv[2] 
+	''' '../../Data/AntTestData.txt' '''
+	outputPath = check_dir(sys.argv[3])
+	ant_names = get_names(namefile)
+	new_dataframe = extract_rows(data, ant_names)
+	new_dataframe.to_csv(sys.argv[3]) 
+	''' ../../Data/processed_ants.csv '''

### 2017-04-13 Scripts/CrunchL.py
commit of scripts

@@ -1,0 +1,38 @@
+#!/usr/bin/env python3
+#Usage: CrunchL.py location_of_files name_of_output_file
+import pandas as pd
+import glob
+import sys
+
+
+def read_table_rb(file):
+	df = pd.read_csv(file, delimiter='\t')
+	mn = df['Likelihood'].mean()
+	print(mn)
+	return(mn)
+
+def read_table_mb(file):
+	df = pd.read_csv(file, delimiter='\t', skiprows=1)
+	return(df["LnL"].mean())
+	
+
+def read_table_rbtl(file):
+	df = pd.read_csv(file, delimiter='\t')
+	return(df["tree_length"].mean())
+
+def read_table_mbtl(file):
+	df = pd.read_csv(file, delimiter='\t', skiprows=1)
+	return(df["TL"].mean())	
+	
+
+if __name__ == "__main__":
+	flist = glob.glob(sys.argv[1])	
+	tuple_list = []
+	for file in flist:
+		print(file)
+		mean_val = read_table_rbtl(file)
+		fname = file.split('_.')[0]
+		tuple_list.append([fname,mean_val])
+	df = pd.DataFrame(tuple_list)
+	print(df)
+	df.to_csv(sys.argv[2], index=False)

### 2017-04-13 Scripts/prune_ants.py

@@ -1,0 +1,49 @@
+#!/usr/bin/env python3
+import dendropy
+import argparse
+import sys
+
+def parse_args(args=None):
+    parser = argparse.ArgumentParser(description='Load & prune trees.')
+    parser.add_argument('File', metavar='FILE', help='File containing trees')
+    parser.add_argument('Burnin', type=int, help='How many trees to discard as burnin')
+    parser.add_argument('Outf', help='Output file name')
+    return parser.parse_args(args)
+    
+def main():
+    retain_list, tl = get_names(file)
+    newtl = tree_ops(tl, burnin, retain_list)
+    print("Writing out pruned trees")
+    newtl.write_to_path(outf, schema='nexus')
+
+def get_names(file):
+    '''Opens trees using Dendropy, and determines which taxa are extinct'''
+    print("Reading in trees")
+    tl = dendropy.TreeList.get_from_path(file, schema="nexus", preserve_underscores=True)
+    print(tl)
+    retain_list = []
+    print("Getting list of taxa to keep")
+    for name in tl.taxon_namespace:
+        if str(name).endswith('X\'', 2) is False:
+            retain_list.append(name)
+    print("To retain:", retain_list)
+    return(retain_list, tl)  
+          
+def tree_ops(tl, burnin, retain_list):
+    '''Extract subtrees of extant taxa'''
+    print("Disgaurding burnin")
+    newtl = tl[burnin:]
+    newTrees = dendropy.TreeList()
+    print("Extracting good taxa")
+    for t in newtl:
+        t = t.extract_tree_with_taxa(retain_list)
+        newTrees.append(t)
+    return(newTrees)
+
+if __name__ == "__main__":
+    args = parse_args()
+    file = args.File
+    burnin = args.Burnin
+    outf = args.Outf
+    
+    main()

### 2017-04-13 Scripts/prune_script.py

@@ -1,0 +1,29 @@
+#!/usr/bin/env python3
+import dendropy
+import argparse
+
+def define_parser():
+	parser = argparse.ArgumentParser(description='Load & prune trees.')
+	parser.add_argument('File', metavar='FILE', help='File containing trees')
+	parser.add_argument('Burnin', type=int help='How many trees to discard as burnin')
+	args = parser.parse_args()
+
+def get_names():
+	tl = dendropy.Tree.get_from_path(file, schema="nexus", preserve_underscores=True)
+	retain_list = []
+	for name in tl.taxon_namespace:
+    	if str(name).endswith('X\'', 2) == False:
+        	retain_list.append(name)
+    return(retain_list)    	
+
+def tree_ops(burnin, retain_list)
+	newtl = tl[burnin:]
+	for t in newtl:
+		t = t.extract_tree_with_taxa(retain_list)
+	return(newtl)
+	
+if __name__ == "__main__":
+	define_parse()
+	retain_list = get_names
+	newtl = tree_ops(burnin, retain_list)
+	newtl.write_to_path('newt', schema='nexus')

### 2017-09-14 Papers/DraftOutline_SSCCM.tex
adding manuscript updates

@@
 	Likewise, the magnitude of change from one state to another may be different between characters, even if the characters mean the same thing (i.e. `0' representing loss and `1' representing gain).
+	For example, in one character, changing from a `0' state to a `1' state could represent a change at a single locus, but at another could represent a much larger change coordinated across many genes.
 	This has complicated the ability to specify $\alpha$ parameters to the Q-matrix \textit{a priori}.
-	This is a stark contrast to molecular data, in which base pairs or nucleotide residues are generally assumed to have similar properties across an alignment. \par
+	This is a stark contrast to molecular data, in which base pairs or amino acid residues are generally assumed to have similar properties across an alignment. \par
 Prior work extending the Mk model has focused on relaxing the assumption of equal character frequencies at stationarity.
-MrBayes implemented a parameter called the Symmetric Dirichlet Hyperprior, which allowed users to place a prior on state frequencies. 
-The probability of observing a change in a character is dependent not only on the probability of change from this character to another, but on the frequency of that character.
+MrBayes implemented a parameter called the Symmetric Dirichlet Hyperprior, which allowed users to place a prior or hyperprior on state frequencies. 
+The probability of observing a change in a character is dependent not only on the probability of change from this character to another, but on the frequency of the starting character.
 Even if a character has a low probability of changing, change in that character may still be observed many times if the stationary frequency of that character is high (i.e., the character is observed many times).
@@
 \par
-This analysis was implemented as a model extension in MrBayes.
-In the case of binary data, this was a Beta prior (the Dirichlet is a generalization of the Beta distribution), which operated in a fashion similar to gamma-distributed rate variation: the distribution is discretized into a user-specified number of categories, the median forward and backward transition rate are calculated for each category, and the likelihood of the character is calculated over each category and summed to form the total likelihood.
+In the case of binary data, the prior used was a Beta prior (the Dirichlet is a generalization of the Beta distribution), which operated in a fashion similar to gamma-distributed rate variation: the distribution is discretized into a user-specified number of categories, the median forward and backward transition rate are calculated for each category, and the likelihood of the character is calculated over each category and summed to form the total likelihood.
 For binary data, the state frequencies are integrated out of the likelihood, making the symmetric Dirichlet a prior.
@@
 However, many datasets that have been analyzed under the CAT model are phylogenomic datasets of thousands of loci, as compared to small morphology datasets.\par
-In RevBayes, we have implemented a number of useful extensions to the Dirichlet prior of RevBayes, and a constrained version of the CAT model. 
+In RevBayes, we have implemented a number of useful extensions to the Dirichlet prior of MrBayes, and a constrained version of the CAT model. 
 Our extension to the symmetric Beta prior allows for the Beta distribution to be asymmetrical.
@@
 The bins draw their stationary frequencies from a Dirichlet to accommodate multistate data.
+
 \subsection{Morphological Phylogeny of the Formicidae}
+
 To test the efficacy of our analytical techniques, we use an ant dataset created from datasets from Barden and Grimaldi and one from Keller.
@@
 As such, they have attracted much work in the world of molecular systematics.
-Ants have also have a rich fossil record, with most subfamilies being represented. \par
+Ants have also have a rich fossil record, with most subfamilies being represented. 
+This fossil record has been used for systematic work, as well as a variety of other ecological and evolutionary questions.\par
 The monophyly of major subfamilies has been consistently supported by most molecular and morphological studies, with the exception of the Cerapachyinae. 
 Molecular systematics had originally shaken up the ant tree of life considerably, breaking up clades considered to be monophyletic based on morphological work.
-However, recent family-wide molecular systematic studies provided support for previous morphological work, supporting a clade of subfamilies Ectatomminae and Heteroponerinae, with Myrmicinae as sister to this clade.
-\par
+For example, based on morphological work, six current subfamilies had been previously considered to be a single subfamily, the Ponerinae, until molecular evidence indicated that two of the clades (Ectatomminae and Heteroponerinae) within that subfamily were demonstrated to be more closely related to other ants on the tree.
+The Ponerinae was then broken into six distinct subfamilies, one of which is called Ponerinae. 
+Recent molecular work has continued to support these six subfamilies as monophyletic, though their relationships to one another remain poorly supported. \par
 Excellent morphological matrices on the ingroup of the Formicidae have been available since the early '90s. 
@@
 Sampling of fossil ants was also expanded to include specimens from the Cretaceous.
-These ants were concluded to be stem lineages, and some could not be concretely placed within lineages. \par
+These ants were concluded to be stem lineages, and some did not demonstrate any particular taxonomic affinity. 
+Even with expanded taxon sampling in the Ponerinae, there is still conflict between the molecular and morphological estimates of phylogenetic relationships. \par
 In this study, we combine the extant matrices of Keller, and the extinct-extant matrix of Barden and Grimaldi to expand the taxon and character sampling.  
@@
 Using this dataset, we strongly support that the use of models that relax key assumptions of the Mk model can greatly improve the fit of the model to the data.
-Some stuff about trees and parameters!
+We also perform simulations to demonstrate that the true number of transition rate asymmetry categories is detectable from the data.
 
@@
 
-	a) Bayes Factor model fitting
-	b) TreeSet viz showing how use of the asymetric beta changes our pattern of exploration of treespace
+\subsection{Data Matrices}
+\subsubsection{Empirical Matrices}
+Several large and well-documented ant matrices were used in this study.
+The first was that of Keller (2011). 
+This matrix is of extant ant groups. 
+It was collected for , and has characters and taxa.
+Of the total character set, were binary and were multistate.
+Because of the scope of Keller's study, the taxon sampling is biased towards and away from . \par
+We also used data from Barden and Grimaldi (2015). 
+This matrix contained both extant and extinct ants, and was collected in order to place a new taxon of ants from the period.
+The matrix contained characters and taxa. \par
 
-Results
+In order to achieve maximum coverage, we merged these character matrices.
+The taxonomic overlap between the two was taxa, and the character overlap (those characters represented in both matrices without any recoding) was .
+The Barden matrix contained characters that were inapplicable to the Keller matrix.
+The Keller matrix contained characters that were innapplicable to the Barden matrix.
+Inapplicable characters were left in the matrix due to their ability to resolve bipartitions in the the authors' respective groups of interest. \par
+Six characters were recoded to bring characters from the original matrices into line with one another.
+Most of these changes simply involved changing the terminology used in reference to the character states.
+For example, Barden and Grimaldi's character 15 is the same as Keller's character 47, but the two matrices had inverse character codings relative to one another.
+The remainder of the changes were changing binary characters to a multistate in cases where one author had more states than the other. \par
+
+	b) Multi-state data matrix
+		b1) Model Fit
+		b2) Topology
+	c) simulated data sets
+		c1) Detectability
+		c2) Correctness
+
+\section{Results}
 
@@
 
-Discussion
+\section{Discussion}
 
@@
 
+\section{Conclusion}
+
+\section{Ackno}
+
 \end{document}

### 2017-09-21 Teaching/my_first_script.py
my first python commit

@@ -1,0 +1,22 @@
+import pandas as pd
+
+def extract_ants(ant_data, query_list):
+	'''Function to get ants out of dataframe'''
+	new_data = ant_data.loc[query_list]
+	return(new_data)
+	
+def get_names(namefile):
+	'''Read in a file of names'''
+	with open(namefile, 'r') as f:
+   		ant_names = f.read().splitlines() 
+	return(ant_names)
+
+def write_out():
+	''' '''
+
+if __name__ == "__main__":
+	ant_data = pd.read_csv("Data/Ants.csv", index_col=0)
+	namefile = "Data/AntNames.txt"
+	query_list = get_names(namefile)
+	extracted_ants = extract_ants(ant_data, query_list)
+	print(extracted_ants)

### 2017-09-21 Papers/DraftOutline_SSCCM.tex

@@
 This matrix is of extant ant groups. 
-It was collected for , and has characters and taxa.
-Of the total character set, were binary and were multistate.
-Because of the scope of Keller's study, the taxon sampling is biased towards and away from . \par
+Keller's matrix was collected with special attention to the poneromorph subgroups (Amblyoponinae, Ectatomminae,
+Heteroponerinae, Paraponerinae, Ponerinae, and Proceratiinae) and has 139 characters and 105 taxa.
+The matrix is highly complete for a morphological dataset, with about 60\% of cells being filled.
+Of the total character set, 100 characters were binary and 39 characters were multistate.
+Because of the scope of Keller's study, the taxon sampling is biased towards the poneromorphs, and away from the other ant subfamilies, including large subfamilies such as the Dolichoderinae and the Formicinae.
+\par
 We also used data from Barden and Grimaldi (2015). 
-This matrix contained both extant and extinct ants, and was collected in order to place a new taxon of ants from the period.
-The matrix contained characters and taxa. \par
-
+This matrix contained both extant and extinct ants, and was collected in order to place stem ants from the Cretaceous period.
+One crown ant amber fossil, \textit{Kyromyrma neffi}, was included in this matrix, but the key feature of this matrix is the sampling of the stem group, including multiple samples from the \textit{Gerontoformica} genus.
+This matrix also expands sampling in the non-ponerine groups that are underrepresented in the Keller matrix.
+The Barden-Grimaldi matrix contained 42 characters and 41 taxa. 
+Of these characters, 26 were binary.\par
 In order to achieve maximum coverage, we merged these character matrices.
-The taxonomic overlap between the two was taxa, and the character overlap (those characters represented in both matrices without any recoding) was .
-The Barden matrix contained characters that were inapplicable to the Keller matrix.
-The Keller matrix contained characters that were innapplicable to the Barden matrix.
+The taxonomic overlap between the two was 11 taxa, and all subfamilies of the Formicidae except the extinct Formiciinae were represented in the matrix. 
+Eleven characters were represented in both matrices without any recoding, and both matrices had many characters that were inapplicable to the other matrices.
 Inapplicable characters were left in the matrix due to their ability to resolve bipartitions in the the authors' respective groups of interest. \par
-Six characters were recoded to bring characters from the original matrices into line with one another.
+Six characters were recoded to make the character states uniform between the two matrices.
 Most of these changes simply involved changing the terminology used in reference to the character states.
 For example, Barden and Grimaldi's character 15 is the same as Keller's character 47, but the two matrices had inverse character codings relative to one another.
-The remainder of the changes were changing binary characters to a multistate in cases where one author had more states than the other. \par
+The remainder of the changes were changing binary characters to a multistate in cases where one author had more states than the other. 
+We will refer to this dataset as the combined dataset. \par
+To test the SHDM model, we used several permutations of these data.
+First, to test the binary version of the model, we used Barden and Grimaldi's dataset with the multistate characters removed.
+This dataset was chosen because the dataset size and completeness is very typical to a morpholgoical dataset, particularly one involving fossils.
+We used this to test the SHDM model.
+We used the combined matrix to also test the full SHDM model.
+Lastly, we used the combined matrix, stripped of multi-state characters to test the binary model on a larger dataset.
+The two datasets including both multistate and binary characters were used to examine the effect of dataset size on the ability to estimate model parameters. \par
 
-	b) Multi-state data matrix
-		b1) Model Fit
-		b2) Topology
-	c) simulated data sets
-		c1) Detectability
-		c2) Correctness
+\subsubsection{Simulated Matrices}
+
+To test the performance of the SHDM model on idealized datasets, we used RevBayes to simulate five sets of 100 replicates of morphological data.
+The first two sets of simulations were based on Barden and Grimaldi's dataset size and tree.
+Under this set of conditions, datasets of 41 taxa and 42 characters were generated in RevBayes.
+In one set, only binary characters were evolved under the Beta model.
+In the other, both binary and mutistate characters were evolved.
+For both datasets, the binary or Dirichlet distribution was discretized into four categories.
+We also created one set of 100 replicates that was generated under a strict Mk model with no state frequency variation to see if the model will overfit state frequency variation where none exists.\par
+The other set of two simulations was based on the full datasets.
+One set of 139 binary characters was generated for the full 105 taxa under the binary model; a second set including multistate characters was generated under the SHDM model.
+The tree used to generate the data was the tree estimated from the combined empirical dataset using the SHDM. \par
+
+\subsection{Phylogenetic Analyses}
+\subsubsection{Empirical Data}
+
+Trees were estimated from each empirical dataset using multiple models of evolution.
+The Mk model was used for each dataset, with an appropriately-sized Q-matrix.
+For the binary-only datasets, the binary model was used.
+The datasets containing both binary and multistate characters were analyzed using the SHDM, with the dimensionality of the Dirichlet process being equal to the largest number of character states. 
+Each dataset was run under several different numbers of discrete state frequency variation categories, from two to eight categories.
+All datasets were corrected for not observing invariant characters. \par
+For the binary datasets, Bayes Factor model fitting was used to assess what the optimal number of discrete categories were.
+We used stepping-stone model fitting to calculate the marginal likelihood.
+We first ran an MCMC analysis to determine how many generations are required to attain convergence (about one million).
+Each of the 50 stepping stones was then run for one million generations.
+Due to computational limits, for the multistate datasets, we used the mean of the MCMC distribution to compare across different discretizations of the model. \par
+
+\subsubsection{Simulated Data}
+
+Because the true tree cannot be known from empirical data, we used simulated data to observe the effect this model has on accuracy of estimation, and to see if it is possible to detect the true number of discrete categories of state frequency variation.
+To this end, for each replicate, all datasets were estimated in RevBayes using the true model.
+Trees were also estimated from all datasets without allowing any state frequency variation (i.e., the Mk model) to simulate the effect of excluding variation (underparameterizing the data) completely.
+Lastly, every dataset was run under the correct model (either the binary model or the SHDM model), but with a misspecified number of categories.
+The number of categories ran from two to eight.
+These categories were chosen because they run from half the true number to double the true number. 
+In total, each dataset was run under one true model and eight misspecified models. \par
 
 \section{Results}
+
+\subsection{Empirical Phylogenetic Analyses}
+\subsection{Binary Data}
+
+The marginal likelihood comparisons for the Barden and Grimaldi dataset, when multistate characters are removed can be seen in Fig. 1a.
+The marginal likelihood comparisons for the combined dataset with the multistate characters removed can be seen in Fig. 1b.
 

### 2018-10-29 Papers/DraftOutline_SSCCM.tex
adding some math

@@
 \documentclass[]{article}
+\usepackage{amsmath}
 
@@
 \section{Methods}
+
+\subsection{Modeling site-heterogeneous state frequencies}
+
+\subsubsection{Discretized beta model}
+We model site-specific state frequencies by assuming the state frequency $\pi_i$ at site $i$ is drawn from a Beta prior distribution with hyperparameters $\alpha,\beta$.
+Then, the likelihood of the data $D$ is computed by marginalizing over the value for $\pi_i$ at each site.
+
+\begin{equation}\label{continuous-likelihood}
+f(D\mid \alpha,\beta) = \prod_{i=1}^n\int_0^1 f(D_i \mid \pi_i)f(\pi_i \mid \alpha,\beta)d\pi_i
+\end{equation}
+
+To simplify the computation, we approximate this integrated likelihood by assuming $\pi_i$ is drawn from a mixture distribution, where the mixture categories are defined deterministically by discretizing a Beta distribution into $k$ bins whose boundaries are defined using $k-1$ quantiles.
+The the value of mixture category $j$ is indicated by $\beta_j$ and is computed as the interquantile median of the $j$-th bin.
+In other words, $\beta_j = I^{-1}_{(2j-1)/2k}(\alpha,\beta)$, where $I_x(\alpha,\beta)$ is the regularized incomplete Beta function.
+The mixing proportion is $1/k$ for each category.
+Then the likelihood is computed by summing over the $k$ discrete mixture categories for $\pi_i$ at each site.
+
+\begin{equation}\label{discretized-likelihood}
+f(D\mid \alpha,\beta) = \prod_{i=1}^n\frac{1}{k}\sum_{j=1}^k f(D_i \mid \pi_i = \beta_j)
+\end{equation}
+
+Note that in the limit as the number of mixture categories approaches infinity, equation \ref{discretized-likelihood} is identical to equation \ref{continuous-likelihood}.
+
+\subsubsection{Site-heterogeneous discrete multistate model}
+
+When analyzing multistate character data with $S$ character states, we assume the state frequencies $\pi_i$ at each site $i$ are drawn from a Dirichlet distribution with concentration parameters $\alpha_1,\ldots,\alpha_S$.
+Again, we use a mixture distribution to simplify the computation, but instead of defining the mixture categories using a discretized Dirichlet distribution, the value $\delta_j$ of mixture category $j$ is assumed to be drawn from the same underlying Dirichlet distribution with parameters $\alpha_1,\ldots,\alpha_S$.
+The mixing proportions are $\theta_j$, which are themselves drawn from a uniform Dirichlet prior distribution.
+
+$$f(D\mid \delta_1,\ldots,\delta_k,\theta_1,\ldots,\theta_k) = \prod_{i=1}^n \sum_{j=1}^k  f(D_i \mid \pi_i = \delta_j)\theta_j$$
 

### 2018-10-29 Papers/DraftOutline_SSCCM.tex
premerge commit

--- 2017-09-21 Papers/DraftOutline_SSCCM.tex
@@
 
+\section{Abstract}
+
+Morphological data is challenging to model due to the arbitrary properties of morphological character codings. 
+The Mk model, a generalization of the Jukes-Cantor model, assumes a that there is an equal rate of forward and backwards transitions in morphological traits.
+In this study, we used an expanded combined set of two ant (Formicidae) data matrices, composed of one extinct-extant matrix and one extant matrix to estimate a phylogeny for the group.
+We use this large matrix to explore methods for relaxing the transition rate symmetry of the Mk model, and introduce the Site Heterogeneous Discrete Morphology (SHDM) to allow equilibrium character frequencies to vary.
+Using a symmetric Dirichlet process as a prior of this modeling allows for the difference in stationary frequency of characters to be expressed more accurately than if expressed with a parsimony method such a generalized Jukes-Cantor model.
+
 \section{Introduction}
@@
 	Recent interest in using Bayesian methods to model morphological evolution to estimate phylogenetic trees has spurred many studies into how well the existing toolkit performs, particularly in relation to parsimony methods.
-	However, there has been little work expanding that toolkit. 
+	However, there has been little work expanding the toolkit of methods available to researchers estimating trees from morphological data. 
 	The most common way to perform likelihood or Bayesian phylogenetic analysis using morphological data is by applying a model of morphological evolution called the Mk model.
@@
 	The Mk model was introduced by Lewis in 2001.
-	As a generalization of the Jukes-Cantor model, it makes the same set of assumptions: that change between any two states is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} states.
-	The model also makes the Markovian assumption that a character can change instantaneously, regardless of previous states. 
+	As a generalization of the Jukes-Cantor model, it makes the same set of assumptions: that change between any two states is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} known states.
+	The model also makes the Markovian assumption that a character can change instantaneously, regardless of previous states, or multiply along a branch. 
 	Any one of these assumptions may strike a reader as problematic for their particular dataset. \par
@@
 \par
-In the case of binary data, the prior used was a Beta prior (the Dirichlet is a generalization of the Beta distribution), which operated in a fashion similar to gamma-distributed rate variation: the distribution is discretized into a user-specified number of categories, the median forward and backward transition rate are calculated for each category, and the likelihood of the character is calculated over each category and summed to form the total likelihood.
-For binary data, the state frequencies are integrated out of the likelihood, making the symmetric Dirichlet a prior.
-For multistate data, the state frequencies are drawn from a Dirichlet distribution.
+In the case of binary data, MrBayes implemented a Beta prior (the Dirichlet is a generalization of the Beta distribution for non-binary data), which operated in a fashion similar to gamma-distributed rate variation: the distribution is discretized into a user-specified number of categories, the median forward and backward transition rate are calculated for each category, and the likelihood of the character is calculated over each category and summed to form the total likelihood.
+For binary data, the state frequencies are integrated out of the likelihood, making the symmetric Dirichlet hyperprior actually a prior.
+For multistate data, the state frequencies are drawn from a Dirichlet distribution, as the Dirichlet is a multi-state generalization of the Beta distribution.
 The parameter was referred to as a hyperprior because users can place a distribution on the parameter to the Beta distribution, and because the state frequencies are not integrated out of the likelihood function for multistate data.
-As implemented in MrBayes, the Beta distribution was assumed to be symmetric, meaning that if there were characters for which the stationary frequency of 0 is high, there would also be corresponding characters for which the stationary frequency of 1 is high.\par
+As implemented in MrBayes, the Beta distribution was assumed to be symmetric, meaning that if there was a class of characters for which the stationary frequency of 0 is high, there would also be corresponding class of characters for which the stationary frequency of 1 is high.\par
 This is a very useful model extension for morphological evolution.
@@
 The CAT model of Lartillot represents one way forward. 
-This model, implemented in PhyloBayes, uses a Dirichlet process to assign individual sites to categories, which differ in their stationary frequencies. 
+This model, implemented in PhyloBayes, uses a Dirichlet process prior to assign individual sites to categories, which differ in their stationary frequencies. 
 In this model, the number of such categories that describe the data is a free parameter. 
 The use of a Dirichlet prior allows for flexibility in the number of states, as opposed to a Beta prior which is limited to binary data.
-However, many datasets that have been analyzed under the CAT model are phylogenomic datasets of thousands of loci, as compared to small morphology datasets.\par
+However, many datasets that have been analyzed under the CAT model are phylogenomic datasets of thousands of loci, as compared to data-limited morphology datasets.\par
 In RevBayes, we have implemented a number of useful extensions to the Dirichlet prior of MrBayes, and a constrained version of the CAT model. 
@@
 This fossil record has been used for systematic work, as well as a variety of other ecological and evolutionary questions.\par
-The monophyly of major subfamilies has been consistently supported by most molecular and morphological studies, with the exception of the Cerapachyinae. 
+The monophyly of major subfamilies has been consistently supported by most molecular studies, with the exception of the Cerapachyinae. 
 Molecular systematics had originally shaken up the ant tree of life considerably, breaking up clades considered to be monophyletic based on morphological work.
 For example, based on morphological work, six current subfamilies had been previously considered to be a single subfamily, the Ponerinae, until molecular evidence indicated that two of the clades (Ectatomminae and Heteroponerinae) within that subfamily were demonstrated to be more closely related to other ants on the tree.
-The Ponerinae was then broken into six distinct subfamilies, one of which is called Ponerinae. 
+The Ponerinae was then broken into six distinct subfamilies, one of which is also called Ponerinae. 
 Recent molecular work has continued to support these six subfamilies as monophyletic, though their relationships to one another remain poorly supported. \par
@@
 Sampling of fossil ants was also expanded to include specimens from the Cretaceous.
-These ants were concluded to be stem lineages, and some did not demonstrate any particular taxonomic affinity. 
+These ants were concluded to be stem lineages, and some did not demonstrate any particular taxonomic affinity based on character data and phylogenetic placement. 
 Even with expanded taxon sampling in the Ponerinae, there is still conflict between the molecular and morphological estimates of phylogenetic relationships. \par
@@
 We will refer to this dataset as the combined dataset. \par
-To test the SHDM model, we used several permutations of these data.
-First, to test the binary version of the model, we used Barden and Grimaldi's dataset with the multistate characters removed.
-This dataset was chosen because the dataset size and completeness is very typical to a morpholgoical dataset, particularly one involving fossils.
-We used this to test the SHDM model.
-We used the combined matrix to also test the full SHDM model.
-Lastly, we used the combined matrix, stripped of multi-state characters to test the binary model on a larger dataset.
-The two datasets including both multistate and binary characters were used to examine the effect of dataset size on the ability to estimate model parameters. \par
+We tested the SHDM model on both binary and binary + multistate datasets.
+First, to test the model on binary-only datasets, we used Barden and Grimaldi's dataset with the multistate characters removed.
+This dataset was chosen because the dataset size and completeness is very typical to a morphological dataset, particularly one involving fossils.
+We also used the combined matrix, stripped of multistate characters to test the model to look at the effect of estimating a larger tree using more characters.
+In both cases, we performed Bayes Factor model fitting to determine how many discrete Beta categories best model the data. \par
+To test the full SHDM model, we used both the Barden-Grimaldi dataset and the combined dataset.
+In testing the SHDM model, we modeled the binary and multistate characters together.
+This allowed us to both examine the impact of allowing character frequencies to be drawn from a Dirichlet, and to examine the effect of adding more data to the phylogenetic question. 
+\par
 
@@
 The first two sets of simulations were based on Barden and Grimaldi's dataset size and tree.
-Under this set of conditions, datasets of 41 taxa and 42 characters were generated in RevBayes.
+Under this set of conditions, datasets of 41 taxa and 42 characters were simulated in RevBayes.
 In one set, only binary characters were evolved under the Beta model.
-In the other, both binary and mutistate characters were evolved.
-For both datasets, the binary or Dirichlet distribution was discretized into four categories.
-We also created one set of 100 replicates that was generated under a strict Mk model with no state frequency variation to see if the model will overfit state frequency variation where none exists.\par
+In the other, both binary and mutistate characters were evolved under the Dirichlet prior.
+For both datasets, the beta or Dirichlet distribution was discretized into four categories.
+\par
 The other set of two simulations was based on the full datasets.
-One set of 139 binary characters was generated for the full 105 taxa under the binary model; a second set including multistate characters was generated under the SHDM model.
+One set of 139 binary characters was generated for the full 105 taxa under the binary model; a second set including multistate characters was generated under the SHDM model, with the distribution discretized into four categories.
 The tree used to generate the data was the tree estimated from the combined empirical dataset using the SHDM. \par
@@
 The datasets containing both binary and multistate characters were analyzed using the SHDM, with the dimensionality of the Dirichlet process being equal to the largest number of character states. 
-Each dataset was run under several different numbers of discrete state frequency variation categories, from two to eight categories.
+Each dataset was run under several different numbers of discrete state frequency variation categories, from two to six categories.
 All datasets were corrected for not observing invariant characters. \par
@@
 We first ran an MCMC analysis to determine how many generations are required to attain convergence (about one million).
-Each of the 50 stepping stones was then run for one million generations.
+Each of the 75 stepping stones was then run for one million generations.
 Due to computational limits, for the multistate datasets, we used the mean of the MCMC distribution to compare across different discretizations of the model. \par
@@
 Because the true tree cannot be known from empirical data, we used simulated data to observe the effect this model has on accuracy of estimation, and to see if it is possible to detect the true number of discrete categories of state frequency variation.
-To this end, for each replicate, all datasets were estimated in RevBayes using the true model.
-Trees were also estimated from all datasets without allowing any state frequency variation (i.e., the Mk model) to simulate the effect of excluding variation (underparameterizing the data) completely.
-Lastly, every dataset was run under the correct model (either the binary model or the SHDM model), but with a misspecified number of categories.
-The number of categories ran from two to eight.
-These categories were chosen because they run from half the true number to double the true number. 
-In total, each dataset was run under one true model and eight misspecified models. \par
+To this end, we estimated trees for each replicate under the true model.
+We also estimated trees for each replicate under 5 misspecified models: the Mk model, 2 and 3 category models (underparameterized models), and 5 and 6 category models (overparameterized models). \par
 
@@
 \subsection{Empirical Phylogenetic Analyses}
-\subsection{Binary Data}
 
@@
 The marginal likelihood comparisons for the combined dataset with the multistate characters removed can be seen in Fig. 1b.
-
+\textbf{Barden Dataset}
 	a) Best fit model?
@@
 	c) Tree set visualization comparisons
+\textbf{Keller Dataset}
+	a) Best fit model?
+	b) Tree comparisons, esp. branch lengths. No asc. bias correction + misspecified model, no correction + misspecified model, correction + correct model, correction + misspecified model 
+	c) Tree set visualization comparisons
+\textbf{Combined Dataset}
+	a) Best fit model?
+
+	b) Tree comparisons, esp. branch lengths. No asc. bias correction + misspecified model, no correction + misspecified model, correction + correct model, correction + misspecified model 
+	c) Tree set visualization comparisons
+
+\subsection{Simulated Phylogenetic Analyses}
+\subsection{Binary 4-cat small}
+
+\subsection{MS4cat small}
+
+\subsection{MS4cat large}
 

### 2018-10-30 Papers/DraftOutline_SSCCM.tex
adding some bibliography stuff

--- 2018-10-29 Papers/DraftOutline_SSCCM.tex
@@
 \usepackage{amsmath} 
+\usepackage{natbib}
 
@@
 
+We model morphological character evolution using the Mk model \citep{lewis2001likelihood}, and additionally account for site-specific heterogeneity in substitution rates among character states by allowing the equilibrium state frequencies $\pi$ to vary among sites, such that each site $i$ has its own state frequency vector $\pi_i$.
+
 \subsubsection{Discretized beta model}
-We model site-specific state frequencies by assuming the state frequency $\pi_i$ at site $i$ is drawn from a Beta prior distribution with hyperparameters $\alpha,\beta$.
+We model site-specific binary state frequencies by assuming the state frequencies $\pi_i$ are drawn from a 2-dimensional Dirichlet (Beta) prior distribution with hyperparameters $\alpha,\beta$.
 Then, the likelihood of the data $D$ is computed by marginalizing over the value for $\pi_i$ at each site.
 
-\begin{equation}\label{continuous-likelihood}
+\begin{equation} \label{continuous-likelihood}
 f(D\mid \alpha,\beta) = \prod_{i=1}^n\int_0^1 f(D_i \mid \pi_i)f(\pi_i \mid \alpha,\beta)d\pi_i
@@
 To simplify the computation, we approximate this integrated likelihood by assuming $\pi_i$ is drawn from a mixture distribution, where the mixture categories are defined deterministically by discretizing a Beta distribution into $k$ bins whose boundaries are defined using $k-1$ quantiles.
-The the value of mixture category $j$ is indicated by $\beta_j$ and is computed as the interquantile median of the $j$-th bin.
-In other words, $\beta_j = I^{-1}_{(2j-1)/2k}(\alpha,\beta)$, where $I_x(\alpha,\beta)$ is the regularized incomplete Beta function.
+The the value of mixture category $j$ is indicated by $\phi_j$ and is computed as the interquantile median of the $j$-th bin.
+In other words, $\phi_j = I^{-1}_{(2j-1)/2k}(\alpha,\beta)$, where $I_x(\alpha,\beta)$ is the regularized incomplete Beta function.
 The mixing proportion is $1/k$ for each category.
@@
 
-\begin{equation}\label{discretized-likelihood}
-f(D\mid \alpha,\beta) = \prod_{i=1}^n\frac{1}{k}\sum_{j=1}^k f(D_i \mid \pi_i = \beta_j)
+\begin{equation} \label{discretized-likelihood}
+f(D\mid \alpha,\beta) = \prod_{i=1}^n\frac{1}{k}\sum_{j=1}^k f(D_i \mid \pi_i = \phi_j)
 \end{equation}
@@
 
-\subsubsection{Site-heterogeneous discrete multistate model}
+\subsubsection{Site-heterogeneous multistate model}
 
@@
 $$f(D\mid \delta_1,\ldots,\delta_k,\theta_1,\ldots,\theta_k) = \prod_{i=1}^n \sum_{j=1}^k  f(D_i \mid \pi_i = \delta_j)\theta_j$$
+
+\subsubsection{Model comparison}
+
+The appropriate number of categories $k$ is selected by comparing estimates of the marginal likelihood under different values for $k$.
+Marginal likelihoods were estimated using the steppingstone sampling method \citep{xie2010improving} implemented in RevBayes.
 
@@
 
+\bibliographystyle{plain}
+\bibliography{bibliography}
+
 \end{document}

### 2018-12-16 Papers/DraftOutline_SSCCM.tex
adding to discussion

@@
 	Making concrete \textit{a priori} statements about the relative probabilities of transitions between character states in a morphological character matrix is challenging. 
-	Unlike a matrix of nucleotide characters, a `0' character at one state may mean that the trait never evolved in one lineage, but was lost in another.
-	This lack of common meaning complicates the interpretation of how change occurs over time.
-	Likewise, the magnitude of change from one state to another may be different between characters, even if the characters mean the same thing (i.e. `0' representing loss and `1' representing gain).
-	For example, in one character, changing from a `0' state to a `1' state could represent a change at a single locus, but at another could represent a much larger change coordinated across many genes.
+	Unlike a matrix of nucleotide characters, in which a state carries common meaning across sites, a `1' state at one character may represent a derived state requiring many underlying genetic changes.
+	At another character, it could represent a derived state requiring only one underlying genetic change.
+	Characters can also differ in meanings with respect to whether they are coded as ancestral or derived.
+	If a matrix was coded with respect to specific outgroups, often the outgroup will mostly have a `0' state, with ingroup taxa having typically having state `1' or higher.
+	This may not be the case for composite matrices or matrices coded without specific reference to an outgroup or character polarities.
+	This lack of common meaning limits a researcher's ability to specify a common mechanism.
 	This has complicated the ability to specify $\alpha$ parameters to the Q-matrix \textit{a priori}.
@@
 
-\subsubsection{Model comparison}
-
-The appropriate number of categories $k$ is selected by comparing estimates of the marginal likelihood under different values for $k$.
-Marginal likelihoods were estimated using the steppingstone sampling method \citep{xie2010improving} implemented in RevBayes.
-
 \subsection{Data Matrices}
@@
 \par
+\subsubsection{Model comparison}
+
+The appropriate number of categories $k$ for each of the two empirical datasets was selected by comparing estimates of the marginal likelihood under different values for $k$.
+Marginal likelihoods were estimated using the steppingstone sampling method \citep{xie2010improving} implemented in RevBayes.
 
@@
 To this end, we estimated trees for each replicate under the true model.
-We also estimated trees for each replicate under 5 misspecified models: the Mk model, 2 and 3 category models (underparameterized models), and 5 and 6 category models (overparameterized models). \par
+We also estimated trees for each replicate under 5 misspecified models: the Mk model, 2 and 3 category models (underparameterized models), and 5 and 6 category models (overparameterized models). 
+Estimation was performed in RevBayes under the generating model.\par
 
@@
 \subsection{Simulated Phylogenetic Analyses}
-\subsection{Binary 4-cat small}
+\subsubsection{Binary Data}
 
-\subsection{MS4cat small}
+In the large datasets, underparameterized models have a deleterious impact on the accuracy of phylogenetic analyses. 
+However, overparameterized models did not show a similar decrease in accuracy.
+In the smaller datasets, accuracy does not seem to be tied to the parameterization of the model. \par
 
-\subsection{MS4cat large}
+\subsubsection{Multistate Data}
+
+In the large, multistate data sets, accuracy is improved using a correctly-specified model over both over- and underparameterized models. 
+In the small datasets, a pattern similar to that of the large binary datasets is observed, in which underparameterization negatively effected accuracy, while overparamterization did not. \par
 
@@
 
-	a) How has the modeling of morphology changed since the Mk was proposed by Lewis? What aspects are comparable to how we model DNA? Which are not? What are potentially interesting future directions? 
-	b) General discussion of usage of treeset visualization
-	c) something else.
+The discussion surrounding the incorporation of morphological data into phylogenetic analyses has largely focused on two endpoints on a spectrum: Parsimony and the Mk model \citep{lewis2001likelihood}.
+The idea behind parsimony is simple: that each character can have its own tree and length (number of steps) on that tree.
+The simplicity of this idea belies the complexity of the implications about character evolution.
+Parsimony can be written out as a likelihood model, referred to as the No Common Mechanisms model.
+However, it has been demonstrated that this model is so complex, and features so many parameters	that it is never chosen by even a liberal information criterion. \par
+In many ways, the Mk model is similar to an unweighted parsimony model, assuming that the transitions between any two states are equally likely.
+But the Mk model assumes a common mechanism across sites, a generalized Jukes-Cantor model.
+Application of among site rate variation (ASRV) across characters allows characters to have differing rates of evolution, accommodating natural variation in phenotypic characters. 
+Allowing the rate of evolution to vary is known to be benefit in phylogenetic analysis \citep{felsenstein78}, and can potentially be more important than an incorrectly-specified model of evolution \citep{lemmon2004importance}.
+Previous work in morphological phylogenetics has also highlighted the importance of appropriate assumptions about rate heterogeneity \citep{Harrison2014}.	\par
+The model discussed here is an effort to create a similar framework to ASRV, but for the actual model of evolution. 
+We may know little about the probability of a particular transition between two character states \textit{a priori}, in the same way that we may know little about the rate of evolution of any particular site.
+Under the discrete Beta model and the SHDM, state frequencies are drawn from either a Beta (binary data) or Dirichlet (multistate data) prior distribution.
+The site likelihoods are then computed by marginalizing over the value of the vector of state frequencies, $\pi_i$.
+Mechanistically, this is very similar to the way in which ASRV is calculated. \par
+In practice, these model allows us to assume multiple possible mechanisms across a single dataset. 
+Previous work on similar models in the software MrBayes \citep{Wright2016} has found that the Mk model is supported using steppingstone model selection in about half of phylogenetic datasets tested.
+However, for modeling the other half, a single common mechanism is not adequate. 
+In this paper, we have implemented the Beta model from MrBayes, and clarified a model for multistate characters to allow modeling of evolution when a single, common mechanism cannot be assumed. \par
+Trees estimated from the empirical data under parsimony are extremely incompletely resolved unless clade constraints are used.
+This is also the case for our model-based estimations. 
+Due to the presence of stem ants the matrices used, particularly those of Barden, would be expected to show multiple models of asymmetric transitions. 
+In some characters, we would expect to see a strong signature of loss: stem ants often display characters similar to that of the wasp outgroup, which crown ants do not.
+We would expect to see these characters lost, but not regained as expected under the Mk model.
+Likewise, due to the presence of ant apomorphies in the crown, we expect to see these characters gained and never lost.
+Together, these character classes argue strongly for a symmetric Beta distribution, in which if there is a set of characters with a certain bias (for example, more `0' to `1' transitions), it is expected that there would be another set of characters with an opposite bias.
+\par
+Indeed, each dataset strongly supports the use of a non-Mk model, with differing numbers of categories (Fig. 1).
+Why then would a model statistically supported to be a better fit to the data not lead to a more resolved tree?
+There are several possible explanations for this.
+First, ants are known to have strong convergence within characters.
+The group is large and diverse, but fairly morphologically constrained.
+While likelihood-based methods are expected to fare better than parsimony when convergence is an issue, they are not perfect and can be mislead, particularly in the presence of model violations such as structured missing data.
+Secondly, the dataset is small relative to the number of taxa.
+It's possible that there simply isn't enough data to address every split in the tree.
+Lastly, I dunno. Morphology what. \par
 
-\section{Conclusion}
+ \par
 

### 2018-12-17 Scripts/TreeComparison.py
treecompare script

@@ -1,61 +1,37 @@
-
-# coding: utf-8
-
-# In[31]:
+#!/anaconda3/bin/python3
 
 import dendropy
 from dendropy.calculate import treecompare
 import itertools
+import glob
+import sys
+import pandas as pd
+
+def read_trees(est_trees, comp_tree):
+	estimated_trees = []
+	list = glob.glob(est_trees)
+	for file in list:
+		estimated_trees.append(dendropy.Tree.get_from_path(file, schema="nexus", preserve_underscores=True, rooting='force-unrooted'))
+	print(estimated_trees[0])
+	comparison_tree = dendropy.Tree.get_from_path(comp_tree, schema="newick", preserve_underscores=True, rooting='force-unrooted', taxon_namespace= estimated_trees[0].taxon_namespace)
+	return(estimated_trees, comparison_tree)
 
 
-# In[163]:
+def perform_comparsions(treelist, comp_tree):
+	df = pd.DataFrame(columns=['RF'])
+	total_diffs = 2*len(comp_tree.nodes())
 
-estimated_trees = []
-mcc = dendropy.Tree.get_from_path("../Data/Morph/output/binary/combined/ants6.mcc.tre", schema="nexus", preserve_underscores=True, rooting='force-unrooted')
-map = dendropy.Tree.get_from_path("../Data/Morph/output/binary/combined/ants6.map.tre", schema="nexus", taxon_namespace= mcc.taxon_namespace, preserve_underscores=True, rooting='force-unrooted')
-con = dendropy.Tree.get_from_path("../Data/Morph/output/binary/combined/ants6.maj.tre", schema="nexus", taxon_namespace= mcc.taxon_namespace, preserve_underscores=True, rooting='force-unrooted')
-allc = dendropy.Tree.get_from_path("../Data/Morph/output/binary/combined/ants6.allcompat.tre", schema="nexus", taxon_namespace= mcc.taxon_namespace, preserve_underscores=True, rooting='force-unrooted')
-
-estimated_trees.append(mcc)
-estimated_trees.append(map)
-estimated_trees.append(con)
-estimated_trees.append(allc)
-
-
-
-# In[164]:
-
-barden_tree = dendropy.Tree.get_from_path("../Data/Morph/Data/Barden.nex", schema="nexus", preserve_underscores=True, rooting='force-unrooted')
-
-
-# In[165]:
-
-include_list = []
-blabels = barden_tree.taxon_namespace.labels()
[…]

### 2018-12-22 Papers/DraftOutline_SSCCM.tex
adding shit for will

@@
 \usepackage{natbib}
-
+\usepackage{graphicx}
+\usepackage{caption}
 \begin{document}
@@
 
-Morphological data is challenging to model due to the arbitrary properties of morphological character codings. 
-The Mk model, a generalization of the Jukes-Cantor model, assumes a that there is an equal rate of forward and backwards transitions in morphological traits.
-In this study, we used an expanded combined set of two ant (Formicidae) data matrices, composed of one extinct-extant matrix and one extant matrix to estimate a phylogeny for the group.
+Morphological data is challenging to model due to the complexity of morphological characters relative to nucleotide sequence or amino acid characters. 
+The Mk model, a generalization of the Jukes-Cantor model, assumes that there is an equal rate of forward and backwards transitions in morphological traits.
+In this study, we use an expanded combined set of two ant (Formicidae) data matrices, composed of one matrix of extinct and extant ants and one matrix of extant ants to estimate a phylogeny for the group.
 We use this large matrix to explore methods for relaxing the transition rate symmetry of the Mk model, and introduce the Site Heterogeneous Discrete Morphology (SHDM) to allow equilibrium character frequencies to vary.
-Using a symmetric Dirichlet process as a prior of this modeling allows for the difference in stationary frequency of characters to be expressed more accurately than if expressed with a parsimony method such a generalized Jukes-Cantor model.
+These two models can be used to explicitly model different character frequencies within one dataset, as opposed to assuming a single common mechanism with respect to among character frequency variation.
 
@@
 	As a generalization of the Jukes-Cantor model, it makes the same set of assumptions: that change between any two states is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} known states.
-	The model also makes the Markovian assumption that a character can change instantaneously, regardless of previous states, or multiply along a branch. 
+	The model also makes the Markovian assumption that a character can change instantaneously, regardless of previous states, and can change multiply along a branch. 
 	Any one of these assumptions may strike a reader as problematic for their particular dataset. \par
@@
 	Unlike a matrix of nucleotide characters, in which a state carries common meaning across sites, a `1' state at one character may represent a derived state requiring many underlying genetic changes.
-	At another character, it could represent a derived state requiring only one underlying genetic change.
+	At another character, it could represent a state requiring only one underlying genetic change.
 	Characters can also differ in meanings with respect to whether they are coded as ancestral or derived.
-	If a matrix was coded with respect to specific outgroups, often the outgroup will mostly have a `0' state, with ingroup taxa having typically having state `1' or higher.
+	If a matrix was coded with respect to specific outgroups, often the outgroup will mostly have a `0' state, with ingroup taxa having typically having state `1' or higher, implying a highly one-directional transition matrix.
 	This may not be the case for composite matrices or matrices coded without specific reference to an outgroup or character polarities.
-	This lack of common meaning limits a researcher's ability to specify a common mechanism.
+	This lack of common meaning limits a researcher's ability to specify a model to apply to the whole matrix.
 	This has complicated the ability to specify $\alpha$ parameters to the Q-matrix \textit{a priori}.
-	This is a stark contrast to molecular data, in which base pairs or amino acid residues are generally assumed to have similar properties across an alignment. \par
+	This is a stark contrast to molecular data, in which base pairs or amino acid residues are generally assumed to have similar properties across an alignment, and exceptions are often defined with respect to known features, such as stem and loop domains.
+\par
 Prior work extending the Mk model has focused on relaxing the assumption of equal character frequencies at stationarity.
 MrBayes implemented a parameter called the Symmetric Dirichlet Hyperprior, which allowed users to place a prior or hyperprior on state frequencies. 
-The probability of observing a change in a character is dependent not only on the probability of change from this character to another, but on the frequency of the starting character.
-Even if a character has a low probability of changing, change in that character may still be observed many times if the stationary frequency of that character is high (i.e., the character is observed many times).
-Likewise, a highly-probable change may be seen relatively rarely if the starting character is observed rarely. 
-Therefore, relaxing the assumption of equal stationary frequencies has been a way of changing the probabilities of observing different changes without making strong \textit{a priori} statements about transition probabilities.
+In most likelihood models of character or sequence evolution, the probability of observing a change in a character is dependent not only on the probability of change from this character to another, but on the frequency of the starting character.
+Even if a character has a low probability of changing, change in that character may still be observed many times if the stationary frequency of that character is high (i.e., the character state is observed many times).
+Likewise, a highly-probable changes may be seen relatively rarely if the starting character is observed rarely. 
+Therefore, relaxing the assumption of equal stationary frequencies has been a way of changing the probabilities of observing different changes without making explicit \textit{a priori} statements about transition probabilities.
 \par
-In the case of binary data, MrBayes implemented a Beta prior (the Dirichlet is a generalization of the Beta distribution for non-binary data), which operated in a fashion similar to gamma-distributed rate variation: the distribution is discretized into a user-specified number of categories, the median forward and backward transition rate are calculated for each category, and the likelihood of the character is calculated over each category and summed to form the total likelihood.
+In the case of binary data, MrBayes implemented a Beta prior (the Dirichlet is a generalization of the Beta distribution for non-binary data), which operated in a fashion similar to gamma-distributed rate variation: the Beta distribution is discretized into a user-specified number of categories, the median forward and backward transition rate are calculated for each category by multiplying together the exchangeability and the character state frequency, and the likelihood of the character is calculated over each category and summed to form the total likelihood.
 For binary data, the state frequencies are integrated out of the likelihood, making the symmetric Dirichlet hyperprior actually a prior.
-For multistate data, the state frequencies are drawn from a Dirichlet distribution, as the Dirichlet is a multi-state generalization of the Beta distribution.
+For multistate data, the state frequencies are fit via MCMC.
 The parameter was referred to as a hyperprior because users can place a distribution on the parameter to the Beta distribution, and because the state frequencies are not integrated out of the likelihood function for multistate data.
-As implemented in MrBayes, the Beta distribution was assumed to be symmetric, meaning that if there was a class of characters for which the stationary frequency of 0 is high, there would also be corresponding class of characters for which the stationary frequency of 1 is high.\par
+As implemented in MrBayes, the Beta distribution was assumed to be symmetric, meaning that if there was a class of characters for which the stationary frequency of 0 is high, there would also be corresponding class of characters for which the stationary frequency of 1 is high.
+A formal description for the mathematics of the model was never published, and some of the mechanisms of operation remain unclear. 
+\par
 This is a very useful model extension for morphological evolution.
@@
 The bins draw their stationary frequencies from a Dirichlet to accommodate multistate data. 
+A mathematical descriptionf of this model can be found in the methods section. \par
 
@@
 Recent molecular work has continued to support these six subfamilies as monophyletic, though their relationships to one another remain poorly supported. \par
-Excellent morphological matrices on the ingroup of the Formicidae have been available since the early '90s. 
+Excellent morphological matrices on the ingroup of the Formicidae have been available since the early '70s. 
 Recent morphological matrices have expanded the sampling of ants from in the large Ponerine family, which was previously undersampled.
@@
 Using this dataset, we strongly support that the use of models that relax key assumptions of the Mk model can greatly improve the fit of the model to the data.
-We also perform simulations to demonstrate that the true number of transition rate asymmetry categories is detectable from the data.
+We also perform simulations to demonstrate that the true number of transition rate asymmetry categories is detectable from the data, and that using an appropriately-specified model of morphological evolution is less likely to lead to incorrect phylogenetic inference than a correctly-specified model.\par
+However, the morphological tree of ants remains poorly-resolved, consistent with prior morphological work in the group.
+We do, however, support the monophyly of what  has been considered the crown group of ants, and support a monophyletic Haidomyrmex ants, as in Barden and Grimaldi (2017). 
+Despite the lack of resolution, we think that this model is more reflective of the evolutionary processes operating within this dataset, and provide a discussion for the continued ambiguity in the group.
+\par
 
@@
 Note that in the limit as the number of mixture categories approaches infinity, equation \ref{discretized-likelihood} is identical to equation \ref{continuous-likelihood}.
+This model is shown as a graphical model in Fig. 1a.
 
@@
 $$f(D\mid \delta_1,\ldots,\delta_k,\theta_1,\ldots,\theta_k) = \prod_{i=1}^n \sum_{j=1}^k  f(D_i \mid \pi_i = \delta_j)\theta_j$$
+
+A graphical model of this model can be seen in Fig. 2B.
 
@@
 We will refer to this dataset as the combined dataset. \par
-We tested the SHDM model on both binary and binary + multistate datasets.
-First, to test the model on binary-only datasets, we used Barden and Grimaldi's dataset with the multistate characters removed.
+We modeled all three datasets according to both models.
+First, to test the Beta model on binary-only datasets, we used Barden and Grimaldi's dataset with the multistate characters removed.
 This dataset was chosen because the dataset size and completeness is very typical to a morphological dataset, particularly one involving fossils.
+Because of the presence of stem ants, we also expect to see strong violations of the assumptions of the Mk model in this dataset.
 We also used the combined matrix, stripped of multistate characters to test the model to look at the effect of estimating a larger tree using more characters.
@@
 In testing the SHDM model, we modeled the binary and multistate characters together.
-This allowed us to both examine the impact of allowing character frequencies to be drawn from a Dirichlet, and to examine the effect of adding more data to the phylogenetic question. 
+This allowed us to both examine the impact of allowing character frequencies to be drawn from a Dirichlet, and to examine the effect of adding more data (both more taxa and more characters) to the phylogenetic question.
 \par
+All estimations were performed in the software RevBayes.
 \subsubsection{Model comparison}
@@
 
-To test the performance of the SHDM model on idealized datasets, we used RevBayes to simulate five sets of 100 replicates of morphological data.
+To test the performance of the SHDM model on idealized datasets, we used RevBayes to simulate four sets of 100 replicates of morphological data.
 The first two sets of simulations were based on Barden and Grimaldi's dataset size and tree.
@@
 In the other, both binary and mutistate characters were evolved under the Dirichlet prior.
-For both datasets, the beta or Dirichlet distribution was discretized into four categories.
+For both datasets, the Beta or Dirichlet distribution was discretized into four categories.
 \par
@@
 We also estimated trees for each replicate under 5 misspecified models: the Mk model, 2 and 3 category models (underparameterized models), and 5 and 6 category models (overparameterized models). 
-Estimation was performed in RevBayes under the generating model.\par
+Estimation was performed in RevBayes.\par
 
@@
 
-The marginal likelihood comparisons for the Barden and Grimaldi dataset, when multistate characters are removed can be seen in Fig. 1a.
-The marginal likelihood comparisons for the combined dataset with the multistate characters removed can be seen in Fig. 1b.
-\textbf{Barden Dataset}
-	a) Best fit model?
-	b) Tree comparisons, esp. branch lengths. No asc. bias correction + misspecified model, no correction + misspecified model, correction + correct model, correction + misspecified model 
-	c) Tree set visualization comparisons
-\textbf{Keller Dataset}
-	a) Best fit model?
-	b) Tree comparisons, esp. branch lengths. No asc. bias correction + misspecified model, no correction + misspecified model, correction + correct model, correction + misspecified model 
-	c) Tree set visualization comparisons
-\textbf{Combined Dataset}
-	a) Best fit model?
-
-	b) Tree comparisons, esp. branch lengths. No asc. bias correction + misspecified model, no correction + misspecified model, correction + correct model, correction + misspecified model 
-	c) Tree set visualization comparisons
+The marginal likelihood comparisons for the Barden and Grimaldi dataset can be seen in Fig. 1a.
+The marginal likelihood comparisons for the combined dataset with the multistate characters removed can be seen in Fig. 1.
+In the Barden dataset, the data support 4 character state frequency variation.
+In the combined dataset, 6 categories are supported for binary data and 7 are supported for multistate. \par
+When a phylogeny is estimated from the Barden dataset under the supported model, it is largely unresolved.
 
@@
 In the large, multistate data sets, accuracy is improved using a correctly-specified model over both over- and underparameterized models. 
-In the small datasets, a pattern similar to that of the large binary datasets is observed, in which underparameterization negatively effected accuracy, while overparamterization did not. \par
+In the small datasets, a pattern similar to that of the large binary datasets is observed, in which underparameterization negatively effected accuracy, while overparameterization did not. \par
 
@@
 The discussion surrounding the incorporation of morphological data into phylogenetic analyses has largely focused on two endpoints on a spectrum: Parsimony and the Mk model \citep{lewis2001likelihood}.
-The idea behind parsimony is simple: that each character can have its own tree and length (number of steps) on that tree.
+The idea behind parsimony is simple: The tree with the least changes implied should be favored.
 The simplicity of this idea belies the complexity of the implications about character evolution.
+Each character can have its own tree and length (number of steps) on that tree.
 Parsimony can be written out as a likelihood model, referred to as the No Common Mechanisms model.
@@
 Previous work in morphological phylogenetics has also highlighted the importance of appropriate assumptions about rate heterogeneity \citep{Harrison2014}.	\par
-The model discussed here is an effort to create a similar framework to ASRV, but for the actual model of evolution. 
+The model discussed here is an effort to apply a similar framework to ASRV, but for the actual model of evolution. 
 We may know little about the probability of a particular transition between two character states \textit{a priori}, in the same way that we may know little about the rate of evolution of any particular site.
@@
 \par
-Indeed, each dataset strongly supports the use of a non-Mk model, with differing numbers of categories (Fig. 1).
+Indeed, each dataset strongly supports the use of a non-Mk model, with differing numbers of categories (Fig. 2).
 Why then would a model statistically supported to be a better fit to the data not lead to a more resolved tree?
-There are several possible explanations for this.
-First, ants are known to have strong convergence within characters.
+The lack of resolution in our recovered topologies is consistent with the lack of resolution in the papers from which the matrices were sampled.
+These matrices were sampled to maximize variation by the original authors. 
+Ants are known to have strong convergence within characters.
 The group is large and diverse, but fairly morphologically constrained.
+The morphological matrices of Barden and Keller were assembled to avoid homoplasious characters, such as sculpturing and body shape. 
+However, in such a large and diverse group, not all homoplasy can be avoided, and previous phylogenetic and taxonomic work in the group has also reported a lack of support for well-supported molecular clades due to a lack of morphological synapomorphies among these groups.
 While likelihood-based methods are expected to fare better than parsimony when convergence is an issue, they are not perfect and can be mislead, particularly in the presence of model violations such as structured missing data.
-Secondly, the dataset is small relative to the number of taxa.
-It's possible that there simply isn't enough data to address every split in the tree.
-Lastly, I dunno. Morphology what. \par
-
+\par
+If, for small matrices, a model that is better fit does not yield a more resolved tree, what does this mean for morphologists? 
+This model is more computationally intense, meaning waiting longer for a result that may not represent a substantial improvement.
+Pragmatically, prior work has supported that model choice is more important when the problem is hard \citep{wright2014, Wright2016}. 
+While a tree may not be more resolved, it is less likely to be incorrect if the model of evolution is not incorrect.
+Even if more resolution cannot be recovered, either due to homoplasy or lack of information, the correct model can help us avoid selecting an incorrect tree.
  \par
 
-\section{Ackno}
+Philosophically, the choice of model should reflect the researcher's knowledge of the data. 
+Models inherently simplify the process of evolution; it is not possible to include every biological process in any model, particularly given the biased sample of the past that we have.
+However, we can hope to increase the availability of models that capture relevant axes of variation. 
+ASRV allows researchers to model different rates of character state change within a dataset.
+Likewise, the discrete Beta model and the SHDM allow researchers to model asymmetrical forward and backward transition rates among characters. 
+In the case of ants, the existence of characters with a strong signature of loss (wasp-like characters) and characters with a stronger signature of gain (ant apomorphies) in the data suggest that it should be expected to find strong support for a model that allows this variation.
+If this is our expectation, then this more complex model should remain in the set of models tested for fit. \par
+The modeling of discrete morphological characters has long been constrained to the end points on a spectrum of possible models. 
+Parsimony represents one extreme in terms of the elaboration of the mechanism. 
+The Mk model represents the other, in its simplicity.
+Researchers may have reasons to believe that either one of these models is a good descriptor of their data, or that assumptions of one model fit better than the other, but that the framework (frequentist or Bayesian) makes more suitable assumptions about the process of evolution.
+Use of new combinations of priors, and borrowing from the molecular toolkit enable researchers to explore more widely the continuum between these two points in a Bayesian context, and to use well-described model fit procedures to test the fit of the model to the data.
+RevBayes enables these sorts of explorations through use of modular design, allowing almost any parameter to be combined with most priors, and including the machinery to instantly simulate under the model.
+The framework enables assumptions about morphological evolution to be advanced, and tested rigorously.
+ \par
 
-\bibliographystyle{plain}
+\section{Conclusions}
+
+WHY WON'T YOU RENDER, THING?
+
+\section{Acknowledgments}
+
+The authors would like to thank Philip Barden, Michael Landis, Jeremy Brown, and Rachel Warnock	 for useful discussion.
+We also thank X, Y and handling editor Z for useful review comments.
+We would also like to thank Christina Kolbmann and Tyler Tran for useful comments on early drafts of the manuscript.
+This work was supported by NSF xxxxxx and LBRN xxxxx to AMW.
+TAH and WCP were supported on NSF XXXXXX to TAH.
+
+\bibliographystyle{sysbio.bst}
 \bibliography{bibliography}
 
+\begin{figure}
+  \caption{A. A graphical model demonstrating the discrete Beta model. B. A graphical model displaying the SHDM. Model realization following Hoehna et al. 2016.}
+    \includegraphics[scale=0.35]{fig/1AB}
+\end{figure}
+
+\begin{figure}
+  \caption{Marginal likelihood calculations for the eight models tested for teach of the four empirical datasets.
+The number of categories into which the Beta (binary data) or Dirichlet (multistate data) is discretized is indicated on the x-axis.
+Note that due to differing dataset sizes, each plot has a unique y-axis.
+    \includegraphics[scale=0.75]{fig/SS_plot_nostar}
+\end{figure}
+
 \end{document}

### 2018-12-23 Papers/DraftOutline_SSCCM.tex
adding last figs

@@
 The number of categories into which the Beta (binary data) or Dirichlet (multistate data) is discretized is indicated on the x-axis.
-Note that due to differing dataset sizes, each plot has a unique y-axis.
-    \includegraphics[scale=0.75]{fig/SS_plot_nostar}
+Note that due to differing dataset sizes, each plot has a unique y-axis.}
+    \includegraphics[scale=0.5]{fig/SS_plot_nostar}
+\end{figure}
+
+\begin{figure}
+  \caption{Accuracy of phylogenetic estimation in large (135 taxa, 163 character) simulated binary datasets under the Beta model, with 4 discrete categories.
+Underparameterized models have a wider spread of accuracy values than appropriately or overparamterized models.}
+    \includegraphics{fig/LargeBinary}
+\end{figure}
+
+\begin{figure}
+  \caption{Accuracy of phylogenetic estimation in small (41 taxa, 26 characters) simulated binary datasets under the Beta model.
+The correct number of simulation categories is 4.
+In datasets this size, there does not seem to be a relationship between parameterization and accuracy.
+}
+    \includegraphics{fig/SmallBinary}
+\end{figure}
+
+\begin{figure}
+  \caption{Accuracy of phylogenetic estimation in small large (135 taxa, 163 character) simulated multistate datasets under the SHDM with four transition rate asymmetry categories.
+Both over- and underparameterization appear to be detrimental to accuracy, though underparamterization is worse.}
+    \includegraphics{fig/LargeMultiState}
+\end{figure}
+
+\begin{figure}
+  \caption{Accuracy of phylogenetic estimation in small (41 taxa, 26 characters) simulated multistate datasets under the SHDM.
+The correct number of simulation categories is 4.
+Much as in the large, binary dataset, underparameterization appears to be more detrimental than overparameterization.}
+    \includegraphics{fig/SmallMS}
 \end{figure}

### 2019-01-03 Papers/DraftOutlineSSCCM.tex
adding results desc of barden tree, fig

--- 2018-12-23 Papers/DraftOutline_SSCCM.tex
@@
 Morphological data is challenging to model due to the complexity of morphological characters relative to nucleotide sequence or amino acid characters. 
-The Mk model, a generalization of the Jukes-Cantor model, assumes that there is an equal rate of forward and backwards transitions in morphological traits.
-In this study, we use an expanded combined set of two ant (Formicidae) data matrices, composed of one matrix of extinct and extant ants and one matrix of extant ants to estimate a phylogeny for the group.
-We use this large matrix to explore methods for relaxing the transition rate symmetry of the Mk model, and introduce the Site Heterogeneous Discrete Morphology (SHDM) to allow equilibrium character frequencies to vary.
+The Mk model is the most commonly-used model for incorporating morphology in maximum likelihood and Bayesian work.
+This model is a generalization of the Jukes-Cantor model, assumes that there is an equal rate of forward and backwards transitions in morphological traits.
+In this study, we explore the use of models that allow asymmetrical transition rates.
+We use an expanded combined set of two ant (Formicidae) data matrices, composed of one matrix of extinct and extant ants and one matrix of extant ants to estimate a phylogeny for the group.
+In particular, we examine the use of two models to allow the equilibrium character frequency to vary. 
+One model, a Beta prior on state frequencies is similar to a model implemented in MrBayes and can be used to allow asymmetrical transition rates in binary characters.
+We also describe a Dirichlet model to allow asymmetrical transitions in multistate characters.
 These two models can be used to explicitly model different character frequencies within one dataset, as opposed to assuming a single common mechanism with respect to among character frequency variation.
@@
 \subsection{Bayesian Modeling of Morphology}
-	Recent interest in using Bayesian methods to model morphological evolution to estimate phylogenetic trees has spurred many studies into how well the existing toolkit performs, particularly in relation to parsimony methods.
+	Recent interest in using Bayesian methods to model morphological evolution to estimate phylogenetic trees has spurred many studies into how well the existing phylogenetic models perform, particularly in relation to parsimony methods.
 	However, there has been little work expanding the toolkit of methods available to researchers estimating trees from morphological data. 
-	The most common way to perform likelihood or Bayesian phylogenetic analysis using morphological data is by applying a model of morphological evolution called the Mk model.
+	The most common way to perform likelihood or Bayesian phylogenetic analysis using discrete morphological data is by applying a model of morphological evolution called the Mk model.
 	Many workers using morphological data have raised questions about the realism of this model.
@@
 	The Mk model was introduced by Lewis in 2001.
-	As a generalization of the Jukes-Cantor model, it makes the same set of assumptions: that change between any two states is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} known states.
+	As a generalization of the Jukes-Cantor model, it makes the same set of assumptions: that change between any two or more states  is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} known states.
 	The model also makes the Markovian assumption that a character can change instantaneously, regardless of previous states, and can change multiply along a branch. 
@@
 	This lack of common meaning limits a researcher's ability to specify a model to apply to the whole matrix.
-	This has complicated the ability to specify $\alpha$ parameters to the Q-matrix \textit{a priori}.
+	This complicates researchers' ability to specify unique $\alpha$ parameters to the Q-matrix, or to apply multiple Q-matrices across the dataset.
+	While the basic model is very simple, some elaborations of the model are common.
+	Assumptions about character evolution, such as ordering, have been common in parsimony analysis, and can be accommodated under the Mk model. 
 	This is a stark contrast to molecular data, in which base pairs or amino acid residues are generally assumed to have similar properties across an alignment, and exceptions are often defined with respect to known features, such as stem and loop domains.
@@
 Even if a character has a low probability of changing, change in that character may still be observed many times if the stationary frequency of that character is high (i.e., the character state is observed many times).
-Likewise, a highly-probable changes may be seen relatively rarely if the starting character is observed rarely. 
+Likewise, a highly probable changes may be seen relatively rarely if the starting character is observed rarely. 
 Therefore, relaxing the assumption of equal stationary frequencies has been a way of changing the probabilities of observing different changes without making explicit \textit{a priori} statements about transition probabilities.
@@
 These ants were concluded to be stem lineages, and some did not demonstrate any particular taxonomic affinity based on character data and phylogenetic placement. 
-Even with expanded taxon sampling in the Ponerinae, there is still conflict between the molecular and morphological estimates of phylogenetic relationships. \par
+Even with expanded taxon sampling in the Ponerinae, there is still conflict between the molecular and morphological estimates of phylogenetic relationships. 
+Absent inputs of additional information, such as clade constraints, the trees are also substantially unresolved. \par
 In this study, we combine the extant matrices of Keller, and the extinct-extant matrix of Barden and Grimaldi to expand the taxon and character sampling.  
@@
 Using this dataset, we strongly support that the use of models that relax key assumptions of the Mk model can greatly improve the fit of the model to the data.
-We also perform simulations to demonstrate that the true number of transition rate asymmetry categories is detectable from the data, and that using an appropriately-specified model of morphological evolution is less likely to lead to incorrect phylogenetic inference than a correctly-specified model.\par
+We also perform simulations to demonstrate that the true number of transition rate asymmetry categories is detectable from the data, and that using an appropriately-specified model of morphological evolution is less likely to lead to incorrect phylogenetic inference than a correctly specified model.\par
 However, the morphological tree of ants remains poorly-resolved, consistent with prior morphological work in the group.
-We do, however, support the monophyly of what  has been considered the crown group of ants, and support a monophyletic Haidomyrmex ants, as in Barden and Grimaldi (2017). 
+We do, however, support the monophyly of what  has been considered the crown group of ants, and support a monophyletic Haidomyrmicine ants, as in Barden and Grimaldi (2017). 
 Despite the lack of resolution, we think that this model is more reflective of the evolutionary processes operating within this dataset, and provide a discussion for the continued ambiguity in the group.
@@
 
-We model morphological character evolution using the Mk model \citep{lewis2001likelihood}, and additionally account for site-specific heterogeneity in substitution rates among character states by allowing the equilibrium state frequencies $\pi$ to vary among sites, such that each site $i$ has its own state frequency vector $\pi_i$.
+We model morphological character evolution using the Mk model \citep{lewis2001likelihood}, and additionally account for site-specific heterogeneity in substitution rates among character states by allowing the equilibrium state frequencies $\pi$ to vary among sites, such that each site $i$ has its own state frequency vector
+$\pi_i$.
 
@@
 
-The marginal likelihood comparisons for the Barden and Grimaldi dataset can be seen in Fig. 1a.
-The marginal likelihood comparisons for the combined dataset with the multistate characters removed can be seen in Fig. 1.
-In the Barden dataset, the data support 4 character state frequency variation.
+The marginal likelihood comparisons for the Barden and Grimaldi dataset can be seen in Fig. 2C and 2D.
+The marginal likelihood comparisons for the combined dataset with the multistate characters removed can be seen in Fig. 2A, and including the multistate characters can be seen in 2B.
+In the Barden datasets, the data support 4 character state frequency variation categories for both the binary and multistate data.
 In the combined dataset, 6 categories are supported for binary data and 7 are supported for multistate. \par
@@
 However, overparameterized models did not show a similar decrease in accuracy.
-In the smaller datasets, accuracy does not seem to be tied to the parameterization of the model. \par
+In the smaller datasets, accuracy does not seem to be tied to the parameterization of the model. 
+Overall, phylogenetic error is lower in larger datasets, even with appropriately parameterized models. \par
 
@@
 In the large, multistate data sets, accuracy is improved using a correctly-specified model over both over- and underparameterized models. 
-In the small datasets, a pattern similar to that of the large binary datasets is observed, in which underparameterization negatively effected accuracy, while overparameterization did not. \par
+In the small datasets, a pattern similar to that of the large binary datasets is observed, in which underparameterization negatively effected accuracy, while overparameterization did not. 
+As with binary data, dataset size has a strong relationship with phylogenetic accuracy. 
+However, the multistate datasets show a stronger relationship between appropriate parameterization and accuracy. 
+This is especially obvious in the small multistate datasets, in which choosing a correct model improves accuracy by approximately 40\%. \par
 
@@
 In this paper, we have implemented the Beta model from MrBayes, and clarified a model for multistate characters to allow modeling of evolution when a single, common mechanism cannot be assumed. \par
+In simulation, for larger datasets, underparameterization of the morphological model appears to have more serious deleterious effects on accuracy than does overparameterization.
+This is expected, as statistical literature indicates that underparameterization often leads to error, but overparameterization often leads to less certainty.
+In the small datasets, we observe different dynamics. 
+The binary datasets do not show a relationship between the appropriateness of the model and accuracy.
+These datasets are quite small - 41 taxa and 26 characters, modeled on Barden and Grimaldi (2015). 
+This dataset is on the extreme small end of the datasets explored by Wright et al. (2016). 
+For the multistate data, the choice of model greatly affects accuracy. 
+Under an overly-simple model of evolution, phylogenetic error is nearly 40\%. 
+But under the correctly-specified model, the error is cut by about 40\%. 
+Why should these two datasets show such differences? 
+It's possible that there is simply not enough information to support a more complex model for the binary data.
+The multistate dataset has nearly twice as many characters for the same amount of taxa. 
+Secondly, with more character states, and more state frequency parameters, the model violations associated with under- and overparameterization are more severe in the multistate datasets.
+\par
 Trees estimated from the empirical data under parsimony are extremely incompletely resolved unless clade constraints are used.
@@
 
-WHY WON'T YOU RENDER, THING?
+Ants have a rich fossil record.
+Due to the presence of stem ants in this fossil record, the assumptions of the Mk model are violated in the data.
+Particularly, due to the loss of wasp-like characters between stem and crown ants, and the gain of ant family and subfamily apomorphies, a single mechanism specifying symmetric transitions is a poor fit to these data.
+Here, we use a discrete Beta model to estimate phylogenetic trees from binary data, and the Site-Heterogeneous Discrete Morphology model to estimate trees from binary and multistate datasets.
+We also use simulation to test these models. 
+We find that using a correctly-specified model can decrease the chances of estimating an incorrect tree.
+Despite these advances, the tree of ants from morphological data remains unresolved.
+However, this model still enables researchers to make assumptions that may strike them as more realistic about their individual datasets.
+\par
 
@@
 
-\bibliographystyle{sysbio.bst}
-\bibliography{bibliography}
+\bibliographystyle{sysbio}
+\bibliography{bibliography.bib}
 

### 2019-01-09 Papers/Draft_SHDM.tex
add changes to text

--- 2019-01-03 Papers/DraftOutlineSSCCM.tex
@@
-\documentclass[]{article}
+\documentclass[]{sysbio}
 \usepackage{amsmath} 
@@
 \begin{document}
-
 \title{Site-Heterogeneous Character Change Models for Morphology}
@@
 Morphological data is challenging to model due to the complexity of morphological characters relative to nucleotide sequence or amino acid characters. 
-The Mk model is the most commonly-used model for incorporating morphology in maximum likelihood and Bayesian work.
-This model is a generalization of the Jukes-Cantor model, assumes that there is an equal rate of forward and backwards transitions in morphological traits.
+Particularly problematic is that morphological characters are coded in a way that is meaningful to human observers, but that this does not necessarily equate to a shared evolutionary mechanism among characters.
+Character states may vary in their meaning between characters, and  the magnitude of underlying genetic changes required to produce a morphological change will also vary among characters.
+The Mk model is the most commonly-used model for incorporating morphology in maximum likelihood and Bayesian phylogenetic estimation.
+This model is a generalization of the Jukes-Cantor model which assumes that there is an equal rate of forward and backwards transitions in morphological traits.
 In this study, we explore the use of models that allow asymmetrical transition rates.
 We use an expanded combined set of two ant (Formicidae) data matrices, composed of one matrix of extinct and extant ants and one matrix of extant ants to estimate a phylogeny for the group.
+There is extensive sampling in the stem ants, which leads to classes of gains and loss within the data, as ant apomorphies are gained, but characters more similar to the outgroup are lost. 
+This violates the assumption of equal transition rates among sites.
 In particular, we examine the use of two models to allow the equilibrium character frequency to vary. 
@@
 \subsection{Bayesian Modeling of Morphology}
-	Recent interest in using Bayesian methods to model morphological evolution to estimate phylogenetic trees has spurred many studies into how well the existing phylogenetic models perform, particularly in relation to parsimony methods.
-	However, there has been little work expanding the toolkit of methods available to researchers estimating trees from morphological data. 
-	The most common way to perform likelihood or Bayesian phylogenetic analysis using discrete morphological data is by applying a model of morphological evolution called the Mk model.
+	Recent interest in using Bayesian methods to model morphological evolution for the estimation of phylogenetic trees has spurred many studies into how well the existing phylogenetic models perform, particularly in relation to parsimony methods.
+	However, there has been little work expanding the toolkit of methods available to researchers estimating trees from morphological data \citep{parins2017, dosreis2018, Alvarez-Carretero441105}. 
+	The most common way to perform likelihood or Bayesian phylogenetic analysis using discrete morphological data is by applying a model of morphological evolution called the Mk model \citep{lewis2001}.
 	Many workers using morphological data have raised questions about the realism of this model.
@@
 	The Mk model was introduced by Lewis in 2001.
-	As a generalization of the Jukes-Cantor model, it makes the same set of assumptions: that change between any two or more states  is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} known states.
+	As a generalization of the Jukes-Cantor model \citep{Jukes1969}, it makes the same set of assumptions: that change between any two or more states  is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} known states.
 	The model also makes the Markovian assumption that a character can change instantaneously, regardless of previous states, and can change multiply along a branch. 
@@
 	Characters can also differ in meanings with respect to whether they are coded as ancestral or derived.
-	If a matrix was coded with respect to specific outgroups, often the outgroup will mostly have a `0' state, with ingroup taxa having typically having state `1' or higher, implying a highly one-directional transition matrix.
+	If a matrix was coded with respect to specific outgroups, characters in the outgroup taxa will mostly have `0' character states, with ingroup taxa having typically having state `1' or higher, implying a highly one-directional transition matrix.
 	This may not be the case for composite matrices or matrices coded without specific reference to an outgroup or character polarities.
-	This lack of common meaning limits a researcher's ability to specify a model to apply to the whole matrix.
-	This complicates researchers' ability to specify unique $\alpha$ parameters to the Q-matrix, or to apply multiple Q-matrices across the dataset.
-	While the basic model is very simple, some elaborations of the model are common.
-	Assumptions about character evolution, such as ordering, have been common in parsimony analysis, and can be accommodated under the Mk model. 
+	This lack of common meaning may strike a researcher as evidence that a single common mechanism model, such as the Mk model, cannot be applied to the dataset.
+	It also complicates researchers' ability to specify unique $\alpha$ parameters to the Q-matrix, or to apply multiple Q-matrices across the dataset, if there is little information by which to model characters separately.
 	This is a stark contrast to molecular data, in which base pairs or amino acid residues are generally assumed to have similar properties across an alignment, and exceptions are often defined with respect to known features, such as stem and loop domains.
 \par
+	While the basic Mk model is very simple, some elaborations of the model are common.
+	Assumptions about character evolution, such as ordering, have been common in parsimony analysis, and can be accommodated under the Mk model. 
+	Models such as a likelihood-based Dollo model have also been implemented. 
+	Morphological models also accommodate among-site rate variation (ASRV), as in molecular data.
+	This is typically achieved via the Gamma distribution, though there is evidence than lognormal distributions may better reflect morphological data, as invariant and very high-rate characters are often excluded from matrices \citep{Harrison2014} \par
 Prior work extending the Mk model has focused on relaxing the assumption of equal character frequencies at stationarity.
 MrBayes implemented a parameter called the Symmetric Dirichlet Hyperprior, which allowed users to place a prior or hyperprior on state frequencies. 
-In most likelihood models of character or sequence evolution, the probability of observing a change in a character is dependent not only on the probability of change from this character to another, but on the frequency of the starting character.
-Even if a character has a low probability of changing, change in that character may still be observed many times if the stationary frequency of that character is high (i.e., the character state is observed many times).
-Likewise, a highly probable changes may be seen relatively rarely if the starting character is observed rarely. 
+In most likelihood models of character or sequence evolution, the probability of observing a change from one character state is dependent not only on the probability of change from this character state to another, but on the frequency of the starting character state.
+Even if a character state  has a low probability of changing to another state (low exchangeability), this change may still be observed many times if the stationary frequency of that character state is high (i.e., the starting character state is observed many times).
+Likewise, a highly probable changes may be seen relatively rarely if the starting character state is observed rarely. 
 Therefore, relaxing the assumption of equal stationary frequencies has been a way of changing the probabilities of observing different changes without making explicit \textit{a priori} statements about transition probabilities.
@@
 However, many datasets that have been analyzed under the CAT model are phylogenomic datasets of thousands of loci, as compared to data-limited morphology datasets.\par
-In RevBayes, we have implemented a number of useful extensions to the Dirichlet prior of MrBayes, and a constrained version of the CAT model. 
+In the software RevBayes, we have implemented a number of useful extensions to the Dirichlet prior of MrBayes, and a constrained version of the CAT model. 
 Our extension to the symmetric Beta prior allows for the Beta distribution to be asymmetrical.
@@
 The bins draw their stationary frequencies from a Dirichlet to accommodate multistate data. 
-A mathematical descriptionf of this model can be found in the methods section. \par
+A mathematical description of this model can be found in the methods section. \par
 
@@
 In this study, we combine the extant matrices of Keller, and the extinct-extant matrix of Barden and Grimaldi to expand the taxon and character sampling.  
-This dataset has interesting properties.
+Barden and Grimaldi's dataset has interesting properties.
 Because there are stem ant lineages represented, there are taxa that have characteristics that are lost after the divergence of the stem lineages.
 This extremely one-sided loss structure violates the Mk model assumption of equal transition rates. 
-We would expect for these characters to be better modeled by a model that can accommodate asymmetrical transition rates.
-Some apomorphies of the ant group are also gained after the divergence of the stem lineages, which also violates the assumption of equal change probabilities.
+We would expect for these characters to be better described by a model that can accommodate asymmetrical transition rates.
+Some apomorphies of the ant group are also gained after the divergence of the stem lineages and not lost frequently, which also violates the assumption of equal change probabilities.
 Because of these model violations, this dataset is an excellent test case for models that relax assumptions of the Mk model. \par
@@
 
-We model morphological character evolution using the Mk model \citep{lewis2001likelihood}, and additionally account for site-specific heterogeneity in substitution rates among character states by allowing the equilibrium state frequencies $\pi$ to vary among sites, such that each site $i$ has its own state frequency vector
-$\pi_i$.
+We model morphological character evolution using the Mk model \citep{lewis2001likelihood}, and additionally account for site-specific heterogeneity in substitution rates among character states by allowing the equilibrium state frequencies $\pi$ to vary among sites, such that each site $i$ has its own state frequency vector $\pi_i$.
 
-\subsubsection{Discretized beta model}
+\subsubsection{Discretized Beta model}
 We model site-specific binary state frequencies by assuming the state frequencies $\pi_i$ are drawn from a 2-dimensional Dirichlet (Beta) prior distribution with hyperparameters $\alpha,\beta$.
@@
 Eleven characters were represented in both matrices without any recoding, and both matrices had many characters that were inapplicable to the other matrices.
-Inapplicable characters were left in the matrix due to their ability to resolve bipartitions in the the authors' respective groups of interest. \par
+Characters that are inapplicable in parts of the tree were retained in the matrix due to their ability to resolve bipartitions in the the authors' respective groups of interest. \par
 Six characters were recoded to make the character states uniform between the two matrices.
@@
 When a phylogeny is estimated from the Barden dataset under the supported model, it is largely unresolved. 
+The tree (Fig. 3) supports the monophyly of crown ants. 
+Haidomyrmicine ants are monophyletic, and less closely related to the crown ants than the Sphecomyrmine ants are, a relationship also supported by Barden and Grimaldi. 
+Many relationships in the crown ants supported by Barden and Grimaldi, such as monophyletic Ponerine, Mymicine, and Formicine clades are also supported in our analysis.  
+We also find support for an assemblage of Doryline, Leptilline and other ants found by the authors, though the exact relationships are different, and poorly-supported. 
+Much as in Barden and Grimaldi's unconstrained analysis, many subfamilies are monophyletic, as expected from molecular analysis, but their relationships to one another remain vague and poorly supported.
+\par
+
+Combined dataset results go here. \par
 
@@
 Indeed, each dataset strongly supports the use of a non-Mk model, with differing numbers of categories (Fig. 2).
-Why then would a model statistically supported to be a better fit to the data not lead to a more resolved tree?
+Even with strongly supported models, the tree is still poorly resolved.
 The lack of resolution in our recovered topologies is consistent with the lack of resolution in the papers from which the matrices were sampled.
@@
 The morphological matrices of Barden and Keller were assembled to avoid homoplasious characters, such as sculpturing and body shape. 
-However, in such a large and diverse group, not all homoplasy can be avoided, and previous phylogenetic and taxonomic work in the group has also reported a lack of support for well-supported molecular clades due to a lack of morphological synapomorphies among these groups.
+However, in such a large and diverse group, not all homoplasy can be avoided, and previous phylogenetic and taxonomic work in the group has also reported a lack of support for well-supported molecular clades due to a dearth of morphological synapomorphies among these groups.
 While likelihood-based methods are expected to fare better than parsimony when convergence is an issue, they are not perfect and can be mislead, particularly in the presence of model violations such as structured missing data.
+However, our methods do reasonably well at teasing apart homoplasy from phylogenetic signal.
+For example, \textit{Martialis Heureka}, an ant that is the only known member of the subfamily Martialinae \citep{rabeling2008}, is supported as sister to \textit{Amblyopone pallipes}, despite elongate mandibles strongly reminiscent of stem ants, such as the Haidonmyrmecine ants.
 \par
-If, for small matrices, a model that is better fit does not yield a more resolved tree, what does this mean for morphologists? 
+If, for small matrices, a model that is better fit does not yield a more resolved tree, or stronger support for the bifurcations that are discovered, what does this mean for morphologists? 
 This model is more computationally intense, meaning waiting longer for a result that may not represent a substantial improvement.
-Pragmatically, prior work has supported that model choice is more important when the problem is hard \citep{wright2014, Wright2016}. 
+Pragmatically, prior work has supported that model choice is more important when the problem is challenging \citep{wright2014, Wright2016}. 
 While a tree may not be more resolved, it is less likely to be incorrect if the model of evolution is not incorrect.
@@
 
-\bibliographystyle{sysbio}
-\bibliography{bibliography.bib}
+\bibliographystyle{sysbio.bst}
+\bibliography{bibliography}
 
@@
   \caption{A. A graphical model demonstrating the discrete Beta model. B. A graphical model displaying the SHDM. Model realization following Hoehna et al. 2016.}
-    \includegraphics[scale=0.35]{fig/1AB}
+    \includegraphics[width=\textwidth]{fig/1AB}
 \end{figure}
@@
 Note that due to differing dataset sizes, each plot has a unique y-axis.}
-    \includegraphics[scale=0.5]{fig/SS_plot_nostar}
+    \includegraphics[width=\textwidth]{fig/SS_plot_nostar}
+\end{figure}
+
+\begin{figure}
+  \caption{Strict consensus tree estimated from the dataset of Barden and Grimaldi (2015) under the SHDM. } 
+    \includegraphics[scale=0.75]{fig/bartree}
 \end{figure}
@@
 \begin{figure}
-  \caption{Accuracy of phylogenetic estimation in small large (135 taxa, 163 character) simulated multistate datasets under the SHDM with four transition rate asymmetry categories.
+  \caption{Accuracy of phylogenetic estimation in sma	ll large (135 taxa, 163 character) simulated multistate datasets under the SHDM with four transition rate asymmetry categories.
 Both over- and underparameterization appear to be detrimental to accuracy, though underparamterization is worse.}
